---
title: "01_CRDS"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(tidymodels)
library(lubridate)
library(fluxible)
```

## Import CRDS data

```{r}
# define function for reading data
read_licor7810 <- function(filename) {
  df = read_delim(filename, skip = 5)
  units = df[1,] |> unlist() |> as.character()
  colnames(df) = paste0(colnames(df), "_", units)
  df = df[-1,]
  return(df)
}

# list file paths
file_paths <- list.files("data/LiCOR/", pattern = "*.data",full.names = TRUE)

# map our function across all elements in the list
df_list <- map(file_paths, read_licor7810)

# concatenate dfs together
crds <- bind_rows(df_list) |> 
  mutate(across(everything(), ~ type.convert(as.character(.), as.is = TRUE))) |> 
  mutate(datetime = ymd_hms(paste(DATE_date, TIME_time))) |> 
  mutate(rn = row_number()) |> 
  mutate(sample_id = NA)
```

## Import metadata

```{r}
metadata <- readxl::read_excel("data/LiCOR/LiCOR_sampling_metadata.xlsx") |> 
  mutate(
    datetime_start = ymd_hms(paste(date, time_start)),
    datetime_end = ymd_hms(paste(date, time_end))
    )

sample_info <- metadata |> 
  select(sample_id, site, condition, time_after_rain, temp_start, temp_end) |> 
  mutate(temp_midpoint = case_when(
    temp_start & temp_end ~ (temp_start + temp_end) / 2,
    temp_start & is.na(temp_end) ~ temp_start,
    temp_end & is.na(temp_start) ~ temp_end
  ))
```

## join metadata with a loop

(I know I'm not supposed to loop in R but jesus christ it's so much easier for this)

(Would love for someone to prove me wrong)

```{r}

# loop over each row
for (i in seq_len(nrow(metadata))) {
  print(i)
  start_time <- metadata$datetime_start[i]
  end_time <- metadata$datetime_end[i]
  current_sample_id <- metadata$sample_id[i]
  
  # find rows where dt is within the current interval
  in_range <- crds$datetime >= start_time & crds$datetime <= end_time
  
  crds$sample_id[in_range] <- current_sample_id
  print(current_sample_id)

}
```

# CH4

plot the flux curves for ch4

```{r}
crds_samples <- crds |> filter(!is.na(sample_id)) |> 
  # remove this weird column
  select(-rn) |>
  # keep only distinct rows (duplicates due to rn column)
  distinct() |>
  # for each measurement ...
  group_by(sample_id) |> 
  arrange(datetime) |> #arrange by datetime in case any data is out of order
  # calcualte time change in seconds
  mutate(dt_start = min(datetime), # noting the start of the measurement
         difftime_sec = as.numeric(datetime - dt_start)
         ) |> 
  ungroup()

crds_samples |> 
  ggplot(
    aes(
      x = difftime_sec,
      y = CH4_ppb
    )
  ) +
  ggpmisc::stat_poly_eq(ggpmisc::use_label("eq")) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

```{r}
# nest the data by sample_id
crds_nested <- crds_samples |> 
  group_by(sample_id) |> 
  nest()
```

```{r}
# define the linear regression model
lm_spec <- linear_reg() |> 
  set_engine("lm")
```

```{r}
# fit the model for each group using map
crds_nested <- crds_nested |> 
  mutate(model = map(data, ~ fit(lm_spec, CH4_ppb ~ difftime_sec, data = .x))) |> 
  # extract model parameters using tidy
  mutate(tidy_model = map(model, ~ tidy(.x$fit)))

# unnest the regression params and 
model_params <- crds_nested |> 
  select(sample_id, tidy_model) |> 
  unnest(tidy_model) |> 
  # pull out slope, remove intercept term b/c we don't want it
  filter(term != "(Intercept)") |> 
  rename(slope = estimate) |> select(-term) |> 
  left_join(sample_info, by = join_by(sample_id))
```

```{r}
crds_samples |> 
  ggplot(
    aes(
      x = datetime, # other option is 'datetime'
      y = CH4_ppb
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_text(data = model_params,
            aes(label = round(slope, 5)), x = 500, y = 2000) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

plot the slopes

```{r}
model_params |> 
  ggplot(
    aes(
      x = time_after_rain,
      y = slope,
      color = temp_midpoint,
    )
  ) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = slope - std.error,
      ymax = slope + std.error
    )
  ) +
  geom_line() +
  facet_wrap(vars(site, condition)) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Calculate flux

1.  Determine linear relationship between concentration (ppbv) and sampling time (minutes). Note that ppm/min equals µL/L/min.
2.  Convert $\alpha_v$ (flux by volume) to $\alpha_m$ (flux by mass) in µg/L/min with ideal gas law

$$
\alpha_m = (\alpha_v \times M \times P) / (R\times T)
$$

where:

-   $\alpha_m$ = µg/L/min

-   $\alpha_v$ = µL/L/min

-   M = molecular weight of target species. In this case, $H_2$ = 2.01568 µg/µmol

-   P = atmospheric pressure (1 atm)

-   R = Gas constant 0.0821 L atm / mol K = 0.0821 µL atm/ µmol K

-   T = temp in K (C + 273)

Then, we calculate flux per unit area:

$$
f_m = (\alpha_m \times V \times 60 \: min/hr) / A
$$

where:

-   $f_m$ = $µg/m^2/hr$

-   $\alpha_m$ defined above

-   V = volume of chamber (L)

-   A = surface area covered by chamber, $m^2$

we can also express in µmol/m2/hr by dividing by CH4 molecular weight

```{r}
min_per_hr = 60
vol_L = 8.2 # 8.2 L -- double check this value!
area_m2 = 0.0312 # chamber area
P_atm = 1 # 1 atm
R = 0.0821 # uL atm / µmol K
ch4_mw = 16.04 # g / mol
temp_K = 273 + 35 # temp placeholder

ch4_flux <- model_params |> 
  mutate(
      ppbv_per_sec = slope,
      ppmv_per_min = ppbv_per_sec / 1000 * 60, # convert to ppm per min
      uL_per_L_per_min = ppmv_per_min,
      ug_per_L_per_min = (uL_per_L_per_min * ch4_mw * P_atm) / (R * temp_K),
      ug_m2_hr = (ug_per_L_per_min * vol_L * min_per_hr) / area_m2
  )
```

Plot

```{r}
p_ch4_flux <- ch4_flux |> 
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
      ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    )
  ) |> 
  filter(time_after_rain != 21) |> 
  ggplot(
    aes(
      x = as.factor(time_after_rain),
      y = ug_m2_hr
    )
  ) +
  geom_col(aes(fill = condition)) +
  scale_fill_manual(values = c("#ffac47", "#4e80c7")) +
  facet_grid(condition ~ site) +
  pilot::theme_pilot() +
  labs(
    x = "Time after rewetting (hrs)",
    y = latex2exp::TeX("$CH_4$ flux $(µg \\; (m^2 \\cdot hr)^{-1})$")
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "none"
  )
p_ch4_flux

ggsave(plot = p_ch4_flux,
       filename = "fig_output/ch4_flux_barchart.png",
       dpi = 1000,
       height = 4,
       width = 5)
```

# CO2

```{r}
crds_samples <- crds |> filter(!is.na(sample_id)) |> 
  # remove this weird column
  select(-rn) |>
  # keep only distinct rows (duplicates due to rn column)
  distinct() |>
  # for each measurement ...
  group_by(sample_id) |> 
  arrange(datetime) |> #arrange by datetime in case any data is out of order
  # calcualte time change in seconds
  mutate(dt_start = min(datetime), # noting the start of the measurement
         difftime_sec = as.numeric(datetime - dt_start)
         ) |> 
  ungroup()

crds_samples |> 
  ggplot(
    aes(
      x = difftime_sec,
      y = CO2_ppm
    )
  ) +
  ggpmisc::stat_poly_eq(ggpmisc::use_label("eq")) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

```{r}
# nest the data by sample_id
crds_nested_co2 <- crds_samples |> 
  group_by(sample_id) |> 
  nest()

# fit the model for each group using map
crds_nested_co2 <- crds_nested_co2 |> 
  mutate(model = map(data, ~ fit(lm_spec, CO2_ppm ~ difftime_sec, data = .x))) |> 
  # extract model parameters using tidy
  mutate(tidy_model = map(model, ~ tidy(.x$fit)))

# unnest the regression params and 
model_params_co2 <- crds_nested_co2 |> 
  select(sample_id, tidy_model) |> 
  unnest(tidy_model) |> 
  # pull out slope, remove intercept term b/c we don't want it
  filter(term != "(Intercept)") |> 
  rename(slope = estimate) |> select(-term) |> 
  left_join(sample_info, by = join_by(sample_id))
```

plot the slopes

```{r}
model_params_co2 |> 
  ggplot(
    aes(
      x = time_after_rain,
      y = slope,
      color = temp_midpoint,
    )
  ) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = slope - std.error,
      ymax = slope + std.error
    )
  ) +
  geom_line() +
  facet_wrap(vars(site, condition)) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## calculate flux

```{r}
min_per_hr = 60
vol_L = 8.2 # 8.2 L -- double check this value!
area_m2 = 0.0312 # chamber area
P_atm = 1 # 1 atm
R = 0.0821 # uL atm / µmol K
co2_mw = 44.01 # g / mol
temp_K = 273 + 35 # temp placeholder

co2_flux <- model_params_co2 |> 
  filter(sample_id != "JCR-SKY-PRE-1") |> 
  filter(sample_id != "JCR-ROS-PRE-1") |> 
  mutate(
      ppmv_sec = slope,
      ppmv_per_min = ppmv_sec * 60, # convert to ppm per min
      uL_per_L_per_min = ppmv_per_min,
      ug_per_L_per_min = (uL_per_L_per_min * co2_mw * P_atm) / (R * temp_K),
      ug_m2_hr = (ug_per_L_per_min * vol_L * min_per_hr) / area_m2,
      umol_m2_hr = ug_m2_hr / co2_mw, # 1 umol co2 is 44.01 ug
      umol_m2_sec = umol_m2_hr / 3600
  )
```

```{r}
p_co2_flux <- co2_flux |> 
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
      ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    )
  ) |> 
  filter(time_after_rain != 21) |>
  #filter(time_after_rain != 6) |>
  ggplot(
    aes(
      x = as.factor(time_after_rain),
      y = umol_m2_sec
    )
  ) +
  #geom_line(color = "gray", linetype = "dotted") +
  geom_col(aes(fill = condition)) +
  # geom_point(
  #   aes(fill = umol_m2_sec),
  #   shape = 21,
  #   color = "black"
  # ) +
  facet_grid(condition ~ site) +
  scale_fill_manual(values = c("#ffac47", "#4e80c7")) +
  pilot::theme_pilot() +
  labs(
    x = "Time after rewetting (hrs)",
    y = latex2exp::TeX("$CO_2$ flux $(µmol \\; (m^2 \\cdot sec)^{-1})$")
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "none"
  )
p_co2_flux

ggsave(
  plot = p_co2_flux,
  filename = "fig_output/co2_flux_barchart.png",
  dpi = 1000,
  height = 4,
  width = 5,
)
```
