---
title: "01_CRDS"
format: html
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(tidymodels)
library(lubridate)
library(fluxible)
```

## Import CRDS data

```{r}
# define function for reading data
read_licor7810 <- function(filename) {
  df = read_delim(filename, skip = 5)
  units = df[1, ] |> unlist() |> as.character()
  colnames(df) = paste0(colnames(df), "_", units)
  df = df[-1, ]
  return(df)
}

# list file paths
file_paths <- list.files("data/LiCOR/", pattern = "*.data", full.names = TRUE)

# map our function across all elements in the list
df_list <- map(file_paths, read_licor7810)

# concatenate dfs together
crds <- bind_rows(df_list) |>
  mutate(across(everything(), ~ type.convert(as.character(.), as.is = TRUE))) |>
  mutate(datetime = ymd_hms(paste(DATE_date, TIME_time))) |>
  mutate(rn = row_number()) |>
  mutate(sample_id = NA)
```

## Import metadata

```{r}
metadata <- readxl::read_excel("data/LiCOR/LiCOR_sampling_metadata.xlsx") |>
  mutate(
    datetime_start = ymd_hms(paste(date, time_start)),
    datetime_end = ymd_hms(paste(date, time_end))
  )

sample_info <- metadata |>
  select(sample_id, site, condition, time_after_rain, temp_start, temp_end) |>
  mutate(
    temp_midpoint = case_when(
      temp_start & temp_end ~ (temp_start + temp_end) / 2,
      temp_start & is.na(temp_end) ~ temp_start,
      temp_end & is.na(temp_start) ~ temp_end
    )
  )
```

## join metadata with a loop

(I know I'm not supposed to loop in R but jesus christ it's so much easier for this)

(Would love for someone to prove me wrong)

```{r}

# loop over each row
for (i in seq_len(nrow(metadata))) {
  print(i)
  start_time <- metadata$datetime_start[i]
  end_time <- metadata$datetime_end[i]
  current_sample_id <- metadata$sample_id[i]

  # find rows where dt is within the current interval
  in_range <- crds$datetime >= start_time & crds$datetime <= end_time

  crds$sample_id[in_range] <- current_sample_id
  print(current_sample_id)
}
```

# CH4

plot the flux curves for ch4

```{r}
crds_samples <- crds |>
  filter(!is.na(sample_id)) |>
  # remove this weird column
  select(-rn) |>
  # keep only distinct rows (duplicates due to rn column)
  distinct() |>
  # for each measurement ...
  group_by(sample_id) |>
  arrange(datetime) |> #arrange by datetime in case any data is out of order
  # calcualte time change in seconds
  mutate(
    dt_start = min(datetime), # noting the start of the measurement
    difftime_sec = as.numeric(datetime - dt_start)
  ) |>
  ungroup()

crds_samples |>
  ggplot(
    aes(
      x = difftime_sec,
      y = CH4_ppb
    )
  ) +
  ggpmisc::stat_poly_eq(ggpmisc::use_label("eq")) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

```{r}
# nest the data by sample_id
crds_nested <- crds_samples |>
  group_by(sample_id) |>
  nest()
```

```{r}
# define the linear regression model
lm_spec <- linear_reg() |>
  set_engine("lm")
```

```{r}
# fit the model for each group using map
crds_nested <- crds_nested |>
  mutate(
    model = map(data, ~ fit(lm_spec, CH4_ppb ~ difftime_sec, data = .x))
  ) |>
  # extract model parameters using tidy
  mutate(tidy_model = map(model, ~ tidy(.x$fit)))

# unnest the regression params and
model_params <- crds_nested |>
  select(sample_id, tidy_model) |>
  unnest(tidy_model) |>
  # pull out slope, remove intercept term b/c we don't want it
  filter(term != "(Intercept)") |>
  rename(slope = estimate) |>
  select(-term) |>
  left_join(sample_info, by = join_by(sample_id))
```

```{r}
crds_samples |>
  ggplot(
    aes(
      x = datetime, # other option is 'datetime'
      y = CH4_ppb
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_text(
    data = model_params,
    aes(label = round(slope, 5)),
    x = 500,
    y = 2000
  ) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

plot the slopes

```{r}
model_params |>
  ggplot(
    aes(
      x = time_after_rain,
      y = slope,
      color = temp_midpoint,
    )
  ) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = slope - std.error,
      ymax = slope + std.error
    )
  ) +
  geom_line() +
  facet_wrap(vars(site, condition)) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Calculate flux

1.  Determine linear relationship between concentration (ppbv) and sampling time (minutes). Note that ppm/min equals µL/L/min.
2.  Convert $\alpha_v$ (flux by volume) to $\alpha_m$ (flux by mass) in µg/L/min with ideal gas law

$$
\alpha_m = (\alpha_v \times M \times P) / (R\times T)
$$

where:

-   $\alpha_m$ = µg/L/min

-   $\alpha_v$ = µL/L/min

-   M = molecular weight of target species. In this case, $H_2$ = 2.01568 µg/µmol

-   P = atmospheric pressure (1 atm)

-   R = Gas constant 0.0821 L atm / mol K = 0.0821 µL atm/ µmol K

-   T = temp in K (C + 273)

Then, we calculate flux per unit area:

$$
f_m = (\alpha_m \times V \times 60 \: min/hr) / A
$$

where:

-   $f_m$ = $µg/m^2/hr$

-   $\alpha_m$ defined above

-   V = volume of chamber (L)

-   A = surface area covered by chamber, $m^2$

we can also express in µmol/m2/hr by dividing by CH4 molecular weight

```{r}
min_per_hr = 60
vol_L = 8.2 # 8.2 L -- double check this value!
area_m2 = 0.0312 # chamber area
P_atm = 1 # 1 atm
R = 0.0821 # uL atm / µmol K
ch4_mw = 16.04 # g / mol
temp_K = 273 + 35 # temp placeholder

ch4_flux <- model_params |>
  mutate(
    ppbv_per_sec = slope,
    ppbv_per_sec_se = std.error,
    ppmv_per_min = ppbv_per_sec / 1000 * 60, # convert to ppm per min
    ppmv_per_min_se = ppbv_per_sec_se / 1000 * 60,
    uL_per_L_per_min = ppmv_per_min,
    uL_per_L_per_min_se = ppmv_per_min_se,
    ug_per_L_per_min = (uL_per_L_per_min * ch4_mw * P_atm) / (R * temp_K),
    ug_per_L_per_min_se = (uL_per_L_per_min_se * ch4_mw * P_atm) / (R * temp_K),
    ug_m2_hr = (ug_per_L_per_min * vol_L * min_per_hr) / area_m2,
    ug_m2_hr_se = (ug_per_L_per_min_se * vol_L * min_per_hr) / area_m2,
    umol_m2_hr = (ug_m2_hr) * (1 / 16.043), # 16.043 ug/umol
    umol_m2_sec = umol_m2_hr / 3600 # 3600 sec per hour
  )
```

## Plot

```{r}
p_ch4_flux <- ch4_flux |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    ),
    time_after_rain = case_when(
      time_after_rain == 0 ~ "Pre",
      TRUE ~ as.factor(time_after_rain)
    ),
    time_after_rain = factor(
      time_after_rain,
      levels = c("Pre", "1", "3", "6", "12", "24", "48")
    )
  ) |>
  filter(time_after_rain != 21) |>
  group_by(site, condition, time_after_rain) |>
  # average replicates if they exist
  dplyr::summarise(
    ug_m2_hr = mean(ug_m2_hr, na.rm = TRUE),
  ) |>
  ungroup() |>
  ggplot(
    aes(
      x = as.factor(time_after_rain),
      y = ug_m2_hr
    )
  ) +
  geom_col(aes(fill = condition)) +
  scale_fill_manual(values = c("#FC9403", "#4E80C7")) +
  facet_grid(site ~ condition) +
  coord_cartesian(ylim = c(0, -42)) +
  theme_minimal() +
  labs(
    x = "Time after rewetting (hrs)",
    y = latex2exp::TeX("$CH_4$ flux $(µg \\; (m^2 \\cdot hr)^{-1})$")
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )
p_ch4_flux

# ggsave(
#   plot = p_ch4_flux,
#   filename = "fig_output/ch4_flux_barchart.png",
#   dpi = 1000,
#   height = 4,
#   width = 5
# )
```

## CH4 in same units as CO2

```{r}
p_ch4_flux2 <- ch4_flux |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    ),
    time_after_rain = case_when(
      time_after_rain == 0 ~ "Pre",
      TRUE ~ as.factor(time_after_rain)
    ),
    time_after_rain = factor(
      time_after_rain,
      levels = c("Pre", "1", "3", "6", "12", "24", "48")
    )
  ) |>
  filter(time_after_rain != 21) |>
  ungroup() |>
  ggplot(
    aes(
      x = as.factor(time_after_rain),
      y = ug_m2_hr
    )
  ) +
  geom_col(aes(fill = condition)) +
  scale_fill_manual(values = c("#FC9403", "#4E80C7")) +
  facet_grid(site ~ condition) +
  coord_cartesian(ylim = c(0, -42)) +
  theme_minimal() +
  labs(
    x = "Time after rewetting (hrs)",
    y = latex2exp::TeX("$CH_4$ flux $(µg \\; (m^2 \\cdot hr)^{-1})$")
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )
p_ch4_flux2
```

# CO2

```{r}
crds_samples <- crds |>
  filter(!is.na(sample_id)) |>
  # remove this weird column
  select(-rn) |>
  # keep only distinct rows (duplicates due to rn column)
  distinct() |>
  # for each measurement ...
  group_by(sample_id) |>
  arrange(datetime) |> #arrange by datetime in case any data is out of order
  # calcualte time change in seconds
  mutate(
    dt_start = min(datetime), # noting the start of the measurement
    difftime_sec = as.numeric(datetime - dt_start)
  ) |>
  ungroup()

crds_samples |>
  ggplot(
    aes(
      x = difftime_sec,
      y = CO2_ppm
    )
  ) +
  ggpmisc::stat_poly_eq(ggpmisc::use_label("eq")) +
  geom_smooth(
    method = "lm",
    formula = "y~x"
  ) +
  geom_line() +
  facet_wrap(vars(sample_id), scales = "free_x")
```

```{r}
# nest the data by sample_id
crds_nested_co2 <- crds_samples |>
  group_by(sample_id) |>
  nest()

# fit the model for each group using map
crds_nested_co2 <- crds_nested_co2 |>
  mutate(
    model = map(data, ~ fit(lm_spec, CO2_ppm ~ difftime_sec, data = .x))
  ) |>
  # extract model parameters using tidy
  mutate(tidy_model = map(model, ~ tidy(.x$fit)))

# unnest the regression params and
model_params_co2 <- crds_nested_co2 |>
  select(sample_id, tidy_model) |>
  unnest(tidy_model) |>
  # pull out slope, remove intercept term b/c we don't want it
  filter(term != "(Intercept)") |>
  rename(slope = estimate) |>
  select(-term) |>
  left_join(sample_info, by = join_by(sample_id))
```

plot the slopes

```{r}
model_params_co2 |>
  ggplot(
    aes(
      x = time_after_rain,
      y = slope,
      color = temp_midpoint,
    )
  ) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = slope - std.error,
      ymax = slope + std.error
    )
  ) +
  geom_line() +
  facet_wrap(vars(site, condition)) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## calculate flux

```{r}
min_per_hr = 60
vol_L = 8.2 # 8.2 L -- double check this value!
area_m2 = 0.0312 # chamber area
P_atm = 1 # 1 atm
R = 0.0821 # uL atm / µmol K
co2_mw = 44.01 # g / mol
temp_K = 273 + 35 # temp placeholder

co2_flux <- model_params_co2 |>
  filter(sample_id != "JCR-SKY-PRE-1") |>
  filter(sample_id != "JCR-ROS-PRE-1") |>
  mutate(
    ppmv_sec = slope,
    ppmv_per_min = ppmv_sec * 60, # convert to ppm per min
    uL_per_L_per_min = ppmv_per_min,
    ug_per_L_per_min = (uL_per_L_per_min * co2_mw * P_atm) / (R * temp_K),
    ug_m2_hr = (ug_per_L_per_min * vol_L * min_per_hr) / area_m2,
    umol_m2_hr = ug_m2_hr / co2_mw, # 1 umol co2 is 44.01 ug
    umol_m2_sec = umol_m2_hr / 3600
  )
```

```{r}
p_co2_flux <- co2_flux |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    ),
    time_after_rain = case_when(
      time_after_rain == 0 ~ "Pre",
      TRUE ~ as.factor(time_after_rain)
    ),
    time_after_rain = factor(
      time_after_rain,
      levels = c("Pre", "1", "3", "6", "12", "24", "48")
    )
  ) |>
  filter(time_after_rain != 21) |>
  ggplot(
    aes(
      x = time_after_rain,
      y = umol_m2_sec
    )
  ) +
  geom_col(aes(fill = condition)) +
  facet_grid(site ~ condition) +
  scale_fill_manual(values = c("#FC9403", "#4E80C7")) +
  theme_minimal() +
  labs(
    x = "Time after rewetting (hrs)",
    y = latex2exp::TeX("$CO_2$ flux $(µmol \\; (m^2 \\cdot sec)^{-1})$")
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )
p_co2_flux

# ggsave(
#   plot = p_co2_flux,
#   filename = "fig_output/co2_flux_barchart.png",
#   dpi = 1000,
#   height = 4,
#   width = 5,
# )
```

# export cowplot of combined fluxes

```{r}
p_combined_fluxes <- cowplot::plot_grid(p_co2_flux, p_ch4_flux, ncol = 2)
p_combined_fluxes

cowplot::save_plot(
  filename = "fig_output/combined_co2_ch4_flux_barplot.pdf",
  plot = p_combined_fluxes,
  base_height = 4,
  base_width = 8
)
```

# Export co2 cache

```{r}
co2_flux |>
  select(
    sample_id,
    site,
    condition,
    time_after_rain,
    umol_m2_sec
  ) |>
  saveRDS(
    file = "cache/co2_flux_umol_m2_sec.rds"
  )
```


# Compare fluxes

ratio of CH4 oxidized to CO2 released. take absolute value of mols methane oxidized over absolute value of mols co2 released

$$
\frac{|CH_4|}{|CO_2|}
$$

```{r}
flux_compare <- co2_flux |>
  mutate(gas = "CO2") |>
  bind_rows(ch4_flux |> mutate(gas = "CH4")) |>
  select(sample_id, site, condition, time_after_rain, gas, umol_m2_sec) |>
  pivot_wider(names_from = gas, values_from = umol_m2_sec) |>
  mutate(ratio_ch4_to_c02 = abs(CH4) / abs(CO2))

#flux_compare |> glimpse()

p_flux_compare <- flux_compare |>
  filter(!str_detect(sample_id, "PRE-1")) |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    ),
    condition = case_when(
      condition == "ROS" ~ "Drought",
      condition == "sky" ~ "No Drought"
    )
  ) |>
  ggplot(
    aes(
      x = time_after_rain,
      y = ratio_ch4_to_c02
    )
  ) +
  geom_smooth(
    method = "loess",
    span = 0.3,
    color = "black",
    alpha = 0.5,
    se = FALSE,
    linetype = "dotted"
  ) +
  geom_point(aes(fill = condition), shape = 21, size = 3) +
  scale_y_continuous(labels = scales::label_number()) +
  scale_fill_manual(values = c("#FC9403", "#4E80C7")) +
  labs(
    x = "Time after rewetting (hours)",
    y = latex2exp::TeX("$\\frac{CH_{4_{ox}}}{CO_{2_{prod}}}$"),
    fill = ""
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    aspect.ratio = 1
  )

p_flux_compare

ggsave(
  filename = "fig_output/flux_compare.pdf",
  plot = p_flux_compare,
  height = 5,
  width = 6
)

# range of ratios?
flux_compare |> pull(ratio_ch4_to_c02) |> min(na.rm = TRUE)
flux_compare |> pull(ratio_ch4_to_c02) |> max(na.rm = TRUE)
flux_compare |> pull(ratio_ch4_to_c02) |> mean(na.rm = TRUE)

# range of CO2 from CH4
# where 50% of C goes to biomass
# expressed in percent
flux_compare |> pull(ratio_ch4_to_c02) |> min(na.rm = TRUE) * 0.5 * 100
flux_compare |> pull(ratio_ch4_to_c02) |> max(na.rm = TRUE) * 0.5 * 100
flux_compare |> pull(ratio_ch4_to_c02) |> mean(na.rm = TRUE) * 0.5 * 100

# how many moles C assimilated into biomass?
# units of umol per m2 per sec (umol_m2_sec)
flux_compare |> pull(CH4) |> min(na.rm = TRUE) * 0.5
flux_compare |> pull(CH4) |> max(na.rm = TRUE) * 0.5
flux_compare |> pull(CH4) |> mean(na.rm = TRUE) |> abs() * 0.5

umol_c_assim_from_ch4_per_m2_per_sec <- flux_compare |>
  pull(CH4) |>
  abs() |>
  max(na.rm = TRUE) * # assume max under dry, aerated conditions
  0.5
umol_c_assim_from_ch4_per_m2_per_sec

# convert to umol per m2 per day
umol_c_assim_from_ch4_per_m2_per_day <- umol_c_assim_from_ch4_per_m2_per_sec *
  86400

umol_c_assim_from_ch4_per_m2_per_day

# convert to umol per g per day
# 1 m2 = 100 cm length * 100 cm width * 10 cm depth = 1E5 cm3
# assume typical soil density of 1 g / cm3
# = 1E5 g soil
umol_c_assim_from_ch4_per_g_per_day <- umol_c_assim_from_ch4_per_m2_per_day *
  (1 / 1E5)
umol_c_assim_from_ch4_per_g_per_day

# convert umol to pmol
pmol_c_assim_from_ch4_per_g_per_day <- umol_c_assim_from_ch4_per_g_per_day * 1E6
pmol_c_assim_from_ch4_per_g_per_day

```