---
title: "02_GC data"
format: html
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(tidymodels)
library(ggpmisc)
library(showtext)
```

```{r}
mtda <- readxl::read_excel("data/trace_gas_metadata.xlsx", skip = 3)

df <- read_delim("data/H2_CO_GC/Dryland_flux_TCaro.csv") |> 
  rename(sample_id = SampleID) |> 
  left_join(mtda, by = join_by(sample_id)) |> 
  arrange(timepoint_hr, dt_min) |> 
  select(sample_id, site, timepoint_hr, dt_min, everything()) |> 
  mutate(dt_min = as.numeric(dt_min)) |> 
  # replace negative or NA values with detection limit of instrument: 0.01 for CO and 0.1 for H2
  mutate(
    H2_ppm = case_when(
      H2_ppm <= 0 | is.na(H2_ppm) ~ 0.1,
      TRUE ~ H2_ppm
    ),
    CO_ppm = case_when(
      CO_ppm <= 0 | is.na(CO_ppm) ~ 0.1,
      TRUE ~ CO_ppm
    )
  )
```

# Take a look at the data

```{r}
df |> 
  filter(site == "JCR") |> 
  ggplot(
    aes(
      x = dt_min,
      y = H2_ppm,
      color = drought
    )
  ) +
    geom_smooth(
      method = "lm",
      se = FALSE
    ) +
    stat_poly_eq(use_label("eq")) +
    geom_point(shape = 1) +
    facet_wrap(vars(timepoint_hr)) +
    theme_bw()
```

# Calculate flux

nest the dataframes

```{r}
tg_nested <- df |> 
  group_by(site, timepoint_hr, drought) |> 
  nest()
```

define the linreg model

```{r}
lm_spec <- linear_reg() |> 
  set_engine("lm")
```

fit the model for each group using `map`

```{r}
tg_flux <- tg_nested |> 
  filter(!is.na(site), !is.na(timepoint_hr)) |> 
  mutate(
    model = map(
      data, ~ fit(
        lm_spec,
        H2_ppm ~ dt_min,
        data = .x
      )
    )
  ) |> 
    # extract model parameters using tidy
  mutate(
    tidy_model = map(model, ~tidy(.x$fit))
  )


  # unnest the regression params
  model_params <- tg_flux |> 
    select(site, tidy_model) |> 
    unnest(tidy_model) |> 
    # extract slope, remove intercept term, we dont want it
    filter(term != "(Intercept)") |> 
    rename(slope = estimate) |> 
    mutate(uL_per_L_per_min = slope) |> 
    select(-term) 
```

## flux calculations

1.  Determine linear relationship between the concentration (ppmv) and the sampling time (minutes). Note that ppmv/min equals µL/L/min.
2.  Convert $\alpha_v$ (flux by volume) to $\alpha_m$ (flux by mass) in µg/L/min with the ideal gas law:

$$
\alpha_m = (\alpha_v \times M \times P) / (R\times T)
$$

where:

-   $\alpha_m$ = µg/L/min

-   $\alpha_v$ = µL/L/min

-   M = molecular weight of target species. In this case, $H_2$ = 2.01568 µg/µmol

-   P = atmospheric pressure (1 atm)

-   R = Gas constant 0.0821 L atm / mol K = 0.0821 µL atm/ µmol K

-   T = temp in K (C + 273)

Then, we calculate flux per unit area:

$$
f_m = (\alpha_m \times V \times 60 \: min/hr) / A
$$

where:

-   $f_m$ = $µg/m^2/hr$

-   $\alpha_m$ defined above

-   V = volume of chamber (L)

-   A = surface area covered by chamber, $m^2$

we can also express in µmol/m2/hr by dividing by H2 molecular weight

```{r}
h2_mw = 2.01568 # ug/umol
P_atm = 1 # 1 atm
R = 0.0821 # uL atm / µmol K
temp_K = 273 + 35 # placeholder temperature!
vol_L = 3.51723 # chamber volume L
min_per_hr = 60 # minutes per hour
area_m2 = 0.0346185 # chamber area m2

h2_flux <- model_params |> 
  mutate(
    ug_per_L_per_min = (uL_per_L_per_min * h2_mw * P_atm) / (R * temp_K),
    ug_m2_hr = (ug_per_L_per_min * vol_L * min_per_hr) / area_m2,
    umol_m2_hr = ug_m2_hr / h2_mw
  )

```

## Plot H2 fluxes

```{r}
h2_flux |> 
  filter(site != "PAL") |> 
  mutate(drought = case_when(drought == "ROS" ~ "Drought", drought == "SKY" ~ "Non-Drought")) |> 
  mutate(site = case_when(
    site == "JCR" ~ "Jalama Canyon Ranch",
    site == "LOM" ~ "Loma Ridge"
  )) |> 
  ggplot(
    aes(
      x = timepoint_hr,
      y = umol_m2_hr,
      color = drought
    )
  ) +
    geom_hline(yintercept = 0, color = "black", linetype = "dotted") +
    geom_smooth(method = "lm", se = FALSE) +
    geom_point(shape = 1) +
    facet_grid(drought ~ site) +
    stat_poly_eq(color = "black", label.x = "right") +
    scale_color_manual(values = c("#a8885e", "#5ea862")) +
    coord_cartesian(xlim = c(0, 48), ylim = c(-20, 10)) +
    theme_pilot() +
    labs(
      x = "Time after rewetting (hr)",
      y = latex2exp::TeX("$µmol \\; (m^{2} \\cdot hr)^{-1}$")
    ) +
    theme(
      legend.position = "none"
    )

```