---
title: "04_IRMS Processing"
format: html
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(ggbiplot)
library(gt)
library(tidyplots)
library(ggpmisc)
library(ComplexHeatmap)
```

# Load GCMS Data


```{r}
gcms <- readxl::read_excel(
  path = "data/metabolomics/mx 884963_Tristan Caro-Karthikeyan_soil extract_primary metabolism_GC-TOF MS_08-2025 submit_tidy.xlsx",
  sheet = "submit",
  skip = 8
) |>
  pivot_longer(
    cols = -c(
      `BinBase name`,
      ret.index,
      `quant mz`,
      `mass spec`,
      PubChem,
      KEGG,
      `InChI Key`
    ),
    names_to = "sample_id",
    values_to = "peak_height"
  ) |>
  mutate(
    site = case_when(
      str_detect(sample_id, "JCR") ~ "JCR",
      str_detect(sample_id, "LOM") ~ "LOM",
    ),
    condition = case_when(
      str_detect(sample_id, "ROS") ~ "Drought",
      str_detect(sample_id, "SKY") ~ "Control"
    ),
    dt = case_when(
      str_detect(sample_id, "72hr") ~ "72hr",
      str_detect(sample_id, "0hr") ~ "0hr",
      str_detect(sample_id, "24hr") ~ "24hr"
    ),
    rep = case_when(
      str_detect(sample_id, "rep1") ~ 1,
      str_detect(sample_id, "rep2") ~ 2,
      str_detect(sample_id, "rep3") ~ 3
    )
  ) |>
  select(sample_id, site, condition, dt, rep, everything()) |>
  rename(compound = `BinBase name`) |>
  # filter out unidentified compounds (represented by integers)
  filter(!str_detect(compound, "^-?\\d+$")) |>
  filter(!sample_id %in% c("BB id", "Pool 1", "Pool 2", "Pool 3", "Pool 4"))

gcms |> glimpse()
gcms |> writexl::write_xlsx("data_output/gcms_data_cleaned.xlsx")
```

```{r}
# inspect all compounds
gcms |> pull(compound) |> unique() |> sort()
```

## Tile plot

```{r}
gcms |>
  ggplot(
    aes(
      x = sample_id,
      y = compound,
      fill = peak_height
    )
  ) +
  geom_tile() +
  scale_fill_viridis_c(trans = "log10") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Convert into matrix for NMDS

```{r}
gcms_nmds <- gcms |>
  select(
    sample_id,
    compound,
    peak_height,
    site,
    condition,
    dt,
    rep
  ) |>
  mutate_at(vars(site, condition, dt, rep), as.factor) |>
  filter(
    !is.na(compound),
    !is.na(site),
    !is.na(condition)
  ) |>
  pivot_wider(
    names_from = sample_id,
    values_from = peak_height
  )

gcms.wide <- gcms |>
  select(
    sample_id,
    site,
    condition,
    dt,
    rep,
    compound,
    peak_height
  ) |>
  filter(!is.na(compound)) |>
  pivot_wider(
    id_cols = c(sample_id, site, condition, dt, rep),
    names_from = compound,
    values_from = peak_height
  )

gcms.pca <- prcomp(
  gcms.wide |>
    # extract the numeric columns for PCA
    select(-c(sample_id, site, condition, dt, rep)) |>
    # missing values are replaced with 0
    mutate(across(everything(), ~ replace_na(.x, 0)))
)

plot(gcms.pca)
```

```{r}
screeplot(gcms.pca, type = "line")

p1 <- ggbiplot(
  gcms.pca,
  groups = gcms.wide$condition,
  var.axes = FALSE
) +
  theme_bw()

p2 <- ggbiplot(
  gcms.pca,
  groups = gcms.wide$site,
  shape = gcms.wide$condition,
  var.axes = FALSE
) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

cowplot::plot_grid(p1, p2, ncol = 2)
```

# GCMS Volcano plotting


```{r}
gcms_volcano <- gcms |>
  group_by(compound) |>
  dplyr::summarize(
    mean_control = mean(peak_height[condition == "Control"], na.rm = TRUE),
    mean_drought = mean(peak_height[condition == "Drought"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[condition == "Control"],
      peak_height[condition == "Drought"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_drought / mean_control),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )

# Create volcano plot
gcms_volcano |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "GC-MS Data",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))

```


```{r}
sig_metabolites <- gcms_volcano |>
  filter(significant == "Significant") |>
  arrange(p_value, log2_fc) |>
  mutate(source = "GCMS")

sig_metabolites |>
  select(compound, p_value, log2_fc) |>
  gt() |>
  data_color(
    columns = c(log2_fc),
    palette = "Blues"
  ) |>
  data_color(columns = c(p_value), palette = "Reds", reverse = TRUE) |>
  fmt_scientific()
```

## export significant metabolite dataframe

```{r}
saveRDS(sig_metabolites, file = "cache/gcms_significant_metabolites.rds")
```

# Load in LCMS datasets and combine with GCMS dataset

Significant metabolites from LCMS
```{r}
# lcms dataset
lcms <- read_rds("cache/lcms_dataset.rds")

# LCMS significant metabolites
lcms_sig <- read_rds("cache/lcms_significant_metabolites.rds") |>
  mutate(source = "LCMS")

overlapping <- intersect(lcms_sig$compound, sig_metabolites$compound)

metab_sig_combined <- lcms_sig |>
  bind_rows(sig_metabolites) |>
  mutate(
    compound_unique = case_when(
      compound %in% overlapping ~ str_c(compound, " (", source, ")"),
      TRUE ~ compound
    )
  )

metab_sig_combined |> arrange(compound)
```

# Examine metabolite patterns from both datasets

```{r}
lcms_volcano <- read_rds("cache/lcms_volcano.rds") |> mutate(source = "LCMS")

overlapping_all <- intersect(lcms_volcano$compound, gcms_volcano$compound)

metab_combined <- lcms_volcano |>
  bind_rows(gcms_volcano |> mutate(source = "GCMS")) |>
  mutate(
    compound_unique = case_when(
      compound %in% overlapping_all ~ str_c(compound, " (", source, ")"),
      TRUE ~ compound
    )
  ) |>
  arrange(compound)

p_metab_combined <- metab_combined |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant,
      label = compound
    )
  ) +
  geom_point(alpha = 0.5, shape = 21) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2)) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_metab_combined
```

```{r}
metab_combined |>
  writexl::write_xlsx(path = "data_output/metab_combined_volcano.xlsx")

metabolite_manual_annotations <- readxl::read_excel(
  "data/metabolomics/metabolite_manual_annotations.xlsx"
)

meta_annot <- metab_combined |>
  left_join(metabolite_manual_annotations, by = "compound") |>
  filter(compound_unique != "mannitol (GCMS)") # remove duplicate
```

```{r}
p_osmo_sig <- meta_annot |>
  # turn depol annotations back to na for this plot
  mutate(
    annotation = case_when(annotation == "depol" ~ NA, TRUE ~ annotation)
  ) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = annotation,
      label = compound
    )
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_point(
    data = function(df) df |> filter(is.na(annotation)),
    shape = 21,
    alpha = 0.1
  ) +
  ggrepel::geom_text_repel(
    data = function(df) {
      df |>
        filter(
          annotation == "osmo",
          #neg_log10_p > -log10(0.05) & abs(log2_fc) > 1
        )
    },
    color = "red",
    size = 3,
    force = 0.5,
    #nudge_y = 3,
    #nudge_x = 5,
    box.padding = 0.3,
    segment.color = "red",
    min.segment.length = 0,
    #segment.curvature = -1e-20,
    segment.alpha = 1,
    alpha = 1,
    ylim = c(-2, Inf),
    xlim = c(-Inf, Inf)
  ) +
  # geom_text(
  #   data = function(df) {
  #     df |>
  #       filter(
  #         annotation == "osmo",
  #         #neg_log10_p > -log10(0.05) & abs(log2_fc) > 1
  #       )
  #   },
  #   color = "red",
  #   hjust = 0,
  #   position = position_nudge(x = 0.1, y = 0.1)
  # ) +
  geom_point(
    data = function(df) df |> filter(annotation == "osmo"),
    shape = 21
  ) +
  scale_fill_manual(values = c("red", "gray")) +
  scale_alpha_manual(values = c(1, 0.2)) +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2), clip = "off") +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_osmo_sig
```

```{r}
p_depol_sig <- meta_annot |>
  filter(!is.na(annotation)) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = annotation,
      label = compound
    )
  ) +
  geom_point(alpha = 0.5, shape = 21) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_fill_manual(values = c("orange", "blue")) +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2)) +
  ggrepel::geom_text_repel(
    data = function(df) {
      df |>
        filter(
          annotation == "depol",
          neg_log10_p > -log10(0.05) & abs(log2_fc) > 1
        )
    },
    color = "black",
    size = 3,
    force = 0.5,
    nudge_y = 1,
    nudge_x = 0,
    box.padding = 1,
    segment.color = "black"
  ) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_depol_sig
```

```{r}
p_combined <- cowplot::plot_grid(
  p_metab_combined,
  p_osmo_sig,
  ncol = 2
)
p_combined

cowplot::save_plot(
  plot = p_combined,
  filename = "fig_output/metabolomics_combined_volcano.pdf",
  base_height = 7,
  base_width = 14
)
```

# import lcms data - Analyze time series data

```{r}
lcms <- read_rds("cache/lcms_full_data.rds")
```


```{r}
annotated_metabolites <- meta_annot |>
  select(compound, annotation) |>
  filter(!is.na(annotation)) |>
  mutate(
    compound = case_when(
      str_detect(compound, "\\(LCMS\\)$") ~ str_remove(
        compound,
        " \\(LCMS\\)$"
      ),
      str_detect(compound, "\\(GCMS\\)$") ~ str_remove(
        compound,
        " \\(GCMS\\)$"
      ),
      TRUE ~ compound
    )
  )

lcms_dt <- annotated_metabolites |>
  left_join(lcms |> rename(compound = metabolite_name), by = "compound")

gcms_dt <- gcms |>
  left_join(annotated_metabolites, by = "compound")
```


```{r}
# define a function for lots of plotting
tidyplot_metabolite_lcms <- function(compound_choice) {
  df <- lcms_dt |>
    mutate(
      site = case_when(
        str_detect(sample_id, "JCR") ~ "Cattle Pasture",
        str_detect(sample_id, "LOM") ~ "Grassland",
      )
    ) |>
    filter(
      !is.na(peak_height),
      !is.na(dt),
      !is.na(site),
      compound == compound_choice
    )

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  if (!any(is.finite(df$peak_height))) {
    return(invisible(NULL))
  }

  df <- df |>
    group_by(site) |>
    mutate(site_max = max(peak_height, na.rm = TRUE)) |>
    ungroup() |>
    filter(is.finite(site_max) & site_max > 0) |>
    group_by(site) |>
    mutate(normalized = peak_height / site_max) |>
    ungroup()

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  df |>
    tidyplot(x = dt, y = normalized, color = condition) |>
    add_mean_bar(alpha = 0.4) |>
    add_sem_errorbar() |>
    add(labs(
      x = "Time after rewetting",
      y = "Normalized Peak Height",
      caption = compound_choice
    )) |>
    add(scale_y_continuous(labels = scales::label_percent(), expand = FALSE)) |>
    adjust_size(height = 60, width = 60) |>
    remove_legend() |>
    split_plot(by = site) |>
    save_plot(
      filename = paste0(
        "fig_output/metabolite_plots/",
        compound_choice,
        "_lcms.png"
      )
    )
}

tidyplot_metabolite_gcms <- function(compound_choice) {
  df <- gcms_dt |>
    mutate(
      site = case_when(
        str_detect(sample_id, "JCR") ~ "Cattle Pasture",
        str_detect(sample_id, "LOM") ~ "Grassland",
      )
    ) |>
    filter(
      !is.na(peak_height),
      !is.na(dt),
      !is.na(site),
      compound == compound_choice
    )

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  if (!any(is.finite(df$peak_height))) {
    return(invisible(NULL))
  }

  df <- df |>
    group_by(site) |>
    mutate(site_max = max(peak_height, na.rm = TRUE)) |>
    ungroup() |>
    filter(is.finite(site_max) & site_max > 0) |>
    group_by(site) |>
    mutate(normalized = peak_height / site_max) |>
    ungroup()

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  df |>
    tidyplot(x = dt, y = normalized, color = condition) |>
    add_mean_bar(alpha = 0.4) |>
    add_sem_errorbar() |>
    add(labs(
      x = "Time after rewetting",
      y = "Normalized Peak Height",
      caption = compound_choice
    )) |>
    add(scale_y_continuous(labels = scales::label_percent(), expand = FALSE)) |>
    adjust_size(height = 60, width = 60) |>
    remove_legend() |>
    split_plot(by = site) |>
    save_plot(
      filename = paste0(
        "fig_output/metabolite_plots/",
        compound_choice,
        "_gcms.png"
      )
    )
}

# uncomment to re-render all bar plots
# for (compound in annotated_metabolites$compound) {
#   tidyplot_metabolite_gcms(compound)
#   tidyplot_metabolite_lcms(compound)
# }
```


```{r}
combined_dt <- gcms_dt |>
  mutate(type = "GCMS") |>
  bind_rows(lcms_dt |> mutate(type = "LCMS")) |>
  select(
    -c(
      PubChem,
      KEGG,
      identifier,
      inchikey,
      m_z,
      rt,
      esi_mode,
      `mass spec`,
      `quant mz`,
      ret.index
    )
  ) |>
  mutate(compound = glue::glue("{compound} ({type})")) |>
  group_by(type, site, condition, compound, dt) |>
  # take mean of replicates at each time point, for each compound, for each site, for each treatment, for each method:
  dplyr::summarize(
    mean_peak_height = mean(peak_height, na.rm = TRUE)
  ) |>
  ungroup() |>
  filter(!is.na(site), !is.na(condition)) |>
  pivot_wider(
    names_from = dt,
    values_from = mean_peak_height
  ) |>
  mutate(
    percent_change_24hr = (`24hr` - `0hr`) / `0hr` * 100,
    percent_change_72hr = (`72hr` - `0hr`) / `0hr` * 100,
    fold_change_24hr = `24hr` / `0hr`,
    fold_change_72hr = `72hr` / `0hr`
  )

combined_dt |> glimpse()

combined_dt |>
  select(
    type,
    site,
    condition,
    compound,
    fold_change_24hr,
    fold_change_72hr,
    percent_change_24hr,
    percent_change_72hr
  ) |>
  arrange(percent_change_72hr) |>
  gt::gt() |>
  gt::data_color(
    columns = percent_change_72hr,
    method = "numeric",
    palette = "viridis"
  ) |>
  gt::fmt_number(decimals = 1)
```

## Analyze timeseries data with volcano plots

```{r}
lcms_volcano_dt <- lcms |>
  group_by(metabolite_name) |>
  # look only at 0hr vs 72hr data
  filter(dt != "24hr") |>
  dplyr::summarize(
    # take average of replicate measurements
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[dt == "0hr"],
      peak_height[dt == "72hr"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_72hr / mean_0hr),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )


# Create volcano plot
lcms_volcano_dt |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "LC-MS Data: 0hr vs 72hr post-rewetting",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))

```

## do the same with GCMS data


```{r}
gcms_volcano_dt <- gcms |>
  group_by(site, compound) |>
  # look only at 0hr vs 72hr data
  filter(dt != "24hr") |>
  dplyr::summarize(
    # take average of replicate measurements
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[dt == "0hr"],
      peak_height[dt == "72hr"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_72hr / mean_0hr),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )


# Create volcano plot
gcms_volcano_dt |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  ggrepel::geom_text_repel(
    data = function(df) df |> filter(significant == "Significant"),
    aes(label = compound),
    color = "black"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  facet_wrap(vars(site)) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "GC-MS Data: 0hr vs 72hr post-rewetting",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))
```

## combined

```{r}
gcms_volcano_dt |>
  bind_rows(lcms_volcano_dt |> rename(compound = metabolite_name)) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change Following Rewetting",
    y = "-Log10 P-Value",
    title = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "None",
    plot.title = element_text(hjust = 0.5)
  )
```




# calculate relative increase and decrease

```{r}

lcms_change <- lcms |>
  rename(compound = metabolite_name) |>
  filter(condition == "Drought") |>
  left_join(metabolite_manual_annotations, by = "compound") |>
  mutate(
    annotation = case_when(annotation == "osmo" ~ "Osmolyte", TRUE ~ NA)
  ) |>
  filter(!is.na(site), !is.na(compound)) |>
  group_by(site, compound) |>
  mutate(
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_24hr = mean(peak_height[dt == "24hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    rel_change_72hr = (mean_72hr - mean_0hr) / mean_0hr,
    rel_change_24hr = (mean_24hr - mean_0hr) / mean_0hr
  ) |>
  ungroup() |>
  group_by(site) |>
  mutate(
    compound = fct_reorder(compound, rel_change_72hr, .desc = TRUE)
  )

gcms_change <- gcms |>
  select(sample_id, site, condition, dt, rep, compound, peak_height) |>
  group_by(site, condition, dt, compound) |>
  summarise(mean_peak_height = mean(peak_height, na.rm = TRUE)) |>
  ungroup() |>
  group_by(site, compound, condition) |>
  mutate(
    # identify the mean peak height at 0hr timepoint
    mean_0hr = mean_peak_height[dt == "0hr"],
    diff_from_t0 = mean_peak_height - mean_0hr,
    rel_change = diff_from_t0 / mean_0hr
  ) |>
  mutate(instrument = "GCMS") |>
  filter(!is.na(compound), !is.na(site), !is.na(condition))

lcms_change <- lcms |>
  select(
    sample_id,
    site,
    condition,
    dt,
    rep,
    compound = metabolite_name,
    peak_height
  ) |>
  group_by(site, condition, dt, compound) |>
  summarise(mean_peak_height = mean(peak_height, na.rm = TRUE)) |>
  ungroup() |>
  group_by(site, compound, condition) |>
  mutate(
    # identify the mean peak height at 0hr timepoint
    mean_0hr = mean_peak_height[dt == "0hr"],
    diff_from_t0 = mean_peak_height - mean_0hr,
    rel_change = diff_from_t0 / mean_0hr
  ) |>
  mutate(instrument = "LCMS") |>
  filter(!is.na(compound))

osmolytes <- annotated_metabolites |>
  filter(annotation == "osmo") |>
  pull(compound)

combined_change <- gcms_change |>
  bind_rows(lcms_change) |>
  mutate(sitecondition = paste(site, condition)) |>
  mutate(osmolyte = compound %in% osmolytes) |>
  group_by(site, condition, compound, dt) |>
  # if there's a duplicated compound, add instrument info
  mutate(
    compound_unique = case_when(
      n() > 1 ~ paste0(compound, " ", instrument),
      TRUE ~ compound
    )
  ) |>
  filter(dt == "72hr") |>
  filter(compound_unique != "mannitol GCMS") |>
  ungroup()
```


```{r}
plot_relative_change <- function(sitecontrol_choice) {
  combined_change |>
    filter(sitecondition == sitecontrol_choice) |>
    group_by(sitecondition) |>
    mutate(
      compound_unique = as.factor(compound_unique),
      compound_unique = fct_reorder(compound_unique, rel_change, .desc = TRUE),
      rel_change_sign = if_else(rel_change >= 0, "pos", "neg")
    ) |>
    ggplot(
      aes(
        x = rel_change,
        y = compound_unique,
        color = osmolyte,
        shape = osmolyte,
        size = osmolyte
      )
    ) +
    geom_vline(xintercept = 0) +
    geom_col(aes(fill = rel_change_sign), color = NA, alpha = 0.1, width = 2) +
    geom_point() +
    ggrepel::geom_text_repel(
      data = function(df) df |> filter(osmolyte),
      aes(label = compound),
      size = 3,
      nudge_x = 0.2,
      nudge_y = 0.1,
      segment.alpha = 0.2
      #position = position_nudge(x = 0.1)
    ) +
    coord_cartesian(xlim = c(-1, 1)) +
    facet_wrap(vars(sitecondition), scales = "free_y") +
    scale_shape_manual(values = c(1, 16)) +
    scale_color_manual(values = c("black", "red")) +
    scale_fill_manual(values = c("#00438d", "#8d0000")) +
    scale_size_manual(values = c(0.5, 2)) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_classic() +
    labs(
      x = "Relative Change\n0 to 72 hrs following rewetting",
      y = ""
    ) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      axis.line.y = element_blank(),
      legend.position = "none"
    )
}

p1 <- plot_relative_change("JCR Control")
p2 <- plot_relative_change("JCR Drought")
p3 <- plot_relative_change("LOM Control")
p4 <- plot_relative_change("LOM Drought")

p1thru4 <- cowplot::plot_grid(p1, p2, p3, p4, ncol = 2)
p1thru4

# cowplot::save_plot(
#   filename = "fig_output/metaB_rel_change.pdf",
#   plot = p1thru4,
#   base_height = 8,
#   base_width = 9
# )
```

## metabolite increase or decrease plot 2

```{r}
timeseries <- gcms_change |>
  bind_rows(lcms_change) |>
  mutate(sitecondition = paste(site, condition)) |>
  mutate(osmolyte = compound %in% osmolytes) |>
  group_by(site, condition, compound, dt) |>
  # if there's a duplicated compound, add instrument info
  mutate(
    compound_unique = case_when(
      n() > 1 ~ paste0(compound, " ", instrument),
      TRUE ~ compound
    )
  ) |>
  filter(compound_unique != "mannitol GCMS") |>
  ungroup() |>
  filter(!is.na(site))

timeseries |> glimpse()
```

```{r}
# weighted mean of all compounds
timeseries_weighted <- timeseries |>
  group_by(site, condition, dt) |>
  summarise(
    rel_change_weighted = sum(rel_change * mean_peak_height, na.rm = TRUE) /
      sum(mean_peak_height, na.rm = TRUE)
  ) |>
  mutate(compound_unique = "weighted_mean", dt = parse_number(dt))

# weighted mean of osmolytes
timeseries_osmo <- timeseries |>
  filter(osmolyte) |>
  group_by(site, condition, dt) |>
  summarise(
    rel_change_weighted = sum(rel_change * mean_peak_height, na.rm = TRUE) /
      sum(mean_peak_height, na.rm = TRUE)
  ) |>
  mutate(compound_unique = "weighted_mean", dt = parse_number(dt))

timeseries |>
  mutate(dt = parse_number(dt)) |>
  ggplot(
    aes(
      x = dt,
      y = rel_change,
      group = compound_unique,
    )
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_line(aes(color = osmolyte, alpha = osmolyte)) +
  geom_point(
    data = function(df) df |> filter(osmolyte),
    color = "black",
    fill = "red",
    shape = 21,
    size = 1
  ) +
  geom_line(
    data = timeseries_weighted,
    aes(y = rel_change_weighted),
    color = "black",
    linewidth = 1
  ) +
  geom_point(
    data = timeseries_weighted,
    aes(y = rel_change_weighted, fill = condition),
    color = "black",
    shape = 21,
    size = 2
  ) +
  geom_text(
    data = function(df) df |> filter(osmolyte, dt == 72),
    aes(label = compound_unique),
    nudge_x = 5,
    size = 2,
    hjust = 0
  ) +
  facet_grid(site ~ condition) +
  coord_cartesian(ylim = c(-1, 4), xlim = c(0, 120)) +
  scale_color_manual(values = c("gray", "red")) +
  scale_fill_manual(values = c("#4E80C7", "#FC9403")) +
  scale_alpha_manual(values = c(0.1, 0.8)) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "Time after rewetting (hr)",
    y = "Metabolite Change"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

```{r}
p5 <- combined_change |>
  filter(osmolyte) |>
  mutate(compound = fct_reorder(compound, rel_change, .desc = TRUE)) |>
  ggplot(aes(
    x = rel_change,
    y = compound,
    color = condition,
    fill = condition
  )) +
  geom_col(position = position_dodge(width = 0.5), width = 0.3) +
  geom_vline(xintercept = 0) +
  facet_wrap(vars(site)) +
  scale_color_manual(values = c("#4E80C7", "#FC9403")) +
  scale_fill_manual(values = c("#4E80C7", "#FC9403")) +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_classic() +
  labs(
    x = "Relative Change\n0 to 72 hrs following rewetting",
    y = ""
  ) +
  theme(
    axis.ticks.y = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.line.y = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.9),
    legend.title = element_blank(),
    legend.background = element_rect(color = "black")
  )
p5

# ggsave(
#   filename = "fig_output/osmolyte_change.pdf",
#   plot = p5,
#   height = 4,
#   width = 8
# )
```

# Export combined dataset

```{r}
lcms |>
  rename(compound = metabolite_name) |>
  bind_rows(gcms) |>
  select(site, condition, dt, rep, compound, peak_height) |>
  writexl::write_xlsx("data_output/metabolomics_export.xlsx")
```

# Calculate alpha diversity

Shannon's:
$$
H = - \Sigma p_i * ln(p_i) \
$$

Simpson's
$$
D = \Sigma (p_i)^2
$$

Simpson's diversity index:
$$
1 - D
$$
```{r}
shannon_diversity <- lcms |>
  rename(compound = metabolite_name) |>
  bind_rows(gcms) |>
  select(site, condition, dt, rep, compound, peak_height) |>
  group_by(site, condition, dt, rep) |>
  mutate(
    total_abundance = sum(peak_height, na.rm = TRUE),
    proportion = peak_height / total_abundance
  ) |>
  filter(proportion > 0, !is.na(site)) |>
  summarise(
    shannon_diversity = -sum(proportion * log(proportion)),
    simpson_diversity = 1 - sum(proportion^2),
    total_compounds = n_distinct(compound),
    pielou_evenness = shannon_diversity / log(total_compounds),
    berger_parker = max(proportion),
    dominant_compound = compound[which.max(proportion)]
  ) |>
  ungroup() |>
  group_by(site, condition, dt) |>
  summarize(
    shannon_mean = mean(shannon_diversity),
    shannon_se = sd(shannon_diversity) / sqrt(n()),
    simpson_mean = mean(simpson_diversity),
    simpson_se = sd(simpson_diversity) / sqrt(n()),
    pielou_mean = mean(pielou_evenness),
    pielou_se = sd(pielou_evenness) / sqrt(n()),
    berger_parker_mean = mean(berger_parker),
    berger_parker_se = sd(berger_parker) / sqrt(n()),
    dominant_compound = if (all(is.na(dominant_compound))) {
      NA_character_
    } else {
      names(sort(table(dominant_compound), decreasing = TRUE))[1]
    }
  )
shannon_diversity |> glimpse()
```

## shannon
```{r}
p_shannon_diversity <- shannon_diversity |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    )
  ) |>
  ggplot(
    aes(
      x = parse_number(dt),
      y = shannon_mean,
      group = condition
    )
  ) +
  # geom_smooth(method = "lm", se = FALSE) +
  # stat_poly_eq(
  #   position = position_dodge(width = 2),
  #   label.x = 0,
  #   label.y = 0,
  #   use_label("eq", "R2")
  # ) +
  geom_line(aes(group = condition), alpha = 0.2) +
  geom_pointrange(
    aes(
      ymin = shannon_mean - shannon_se,
      ymax = shannon_mean + shannon_se,
      shape = site,
      fill = condition
    )
  ) +
  facet_wrap(vars(site), scales = "free_y", ncol = 1) +
  coord_cartesian(ylim = c(3.5, 5.3)) +
  scale_fill_manual(values = c("#4E80C7", "#FC9403")) +
  scale_shape_manual(values = c(21, 22)) +
  labs(
    x = "Time after rewetting (hr)",
    y = "Shannon Diversity"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.background = element_blank(),
    legend.position = "none",
    axis.text = element_text(color = "black", face = "bold")
  )
p_shannon_diversity

ggsave(
  filename = "fig_output/metabolite_shannon.pdf",
  plot = p_shannon_diversity,
  height = 5,
  width = 4
)
```

## simpson 
```{r}
p_simpson_diversity <- shannon_diversity |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    )
  ) |>
  ggplot(
    aes(
      x = parse_number(dt),
      y = simpson_mean
    )
  ) +
  geom_line(aes(group = condition), alpha = 0.2) +
  geom_pointrange(
    aes(
      ymin = simpson_mean - simpson_se,
      ymax = simpson_mean + simpson_se,
      shape = site,
      fill = condition
    )
  ) +
  facet_wrap(vars(site)) +
  #coord_cartesian(ylim = c(3.5, 5.3)) +
  scale_fill_manual(values = c("#4E80C7", "#FC9403")) +
  scale_shape_manual(values = c(21, 22)) +
  labs(
    x = "Time after rewetting (hr)",
    y = "Simpson Diversity"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.background = element_blank(),
    legend.position = "none",
    axis.text = element_text(color = "black", face = "bold")
  )
p_simpson_diversity

# ggsave(
#   filename = "fig_output/metabolite_simpson.pdf",
#   plot = p_simpson_diversity,
#   height = 4,
#   width = 5
# )
```

## berger_parker

```{r}
p_bp_dominance <- shannon_diversity |>
  mutate(
    site = case_when(
      site == "JCR" ~ "Cattle Pasture",
      site == "LOM" ~ "Grassland"
    )
  ) |>
  ggplot(
    aes(
      x = parse_number(dt),
      y = berger_parker_mean
    )
  ) +
  geom_line(aes(group = condition), alpha = 0.2) +
  geom_text(
    aes(label = dominant_compound),
    size = 3,
    position = position_nudge(x = 10, y = 0.01)
  ) +
  geom_pointrange(
    aes(
      ymin = berger_parker_mean - berger_parker_se,
      ymax = berger_parker_mean + berger_parker_se,
      shape = site,
      fill = condition
    )
  ) +
  facet_wrap(vars(site)) +
  #coord_cartesian(ylim = c(3.5, 5.3)) +
  scale_fill_manual(values = c("#4E80C7", "#FC9403")) +
  scale_shape_manual(values = c(21, 22)) +
  labs(
    x = "Time after rewetting (hr)",
    y = "Berger-Parker Dominance"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.background = element_blank(),
    legend.position = "none",
    axis.text = element_text(color = "black" face = "bold")
  )
p_bp_dominance

ggsave(
  filename = "fig_output/metabolite_pielou.pdf",
  plot = p_pielou_evennness,
  height = 4,
  width = 5
)
```

```{r}
library(ordr)

# load data
metab <- read_csv("data_output/exported_metaB.csv") |>
  filter(!is.na(compound), !is.na(site), !is.na(dt)) |>
  mutate(dt = fct_rev(factor(parse_number(dt)))) |>
  pivot_wider(
    id_cols = c(site, condition, dt),
    names_from = compound,
    values_from = peak_height
  )
```

# ordinate

```{r}
metab_pca <- ordinate(
  metab,
  cols = -c(site, condition, dt),
  model = ~ prcomp(., scale. = TRUE)
)

metab_pca
```

# screeeeeeee


```{r}
metab_pca |>
  broom::tidy() |>
  print() |>
  ggplot(aes(x = name, y = prop_var)) +
  geom_col() +
  labs(x = "", y = "Proportion of inertia")

```

# biplot


```{r}
ggbiplot(
  metab_pca,
  sec.axes = "cols",
  scale.factor = 2,
  axis.type = "predictive"
) +
  geom_rows_point(aes(color = condition, shape = site), size = 5) +
  scale_color_manual(values = c("#4E80C7", "#FC9403")) +
  geom_line(
    aes(group = interaction(site, condition)),
    arrow = arrow(length = unit(0.1, "inches"), ends = "first"),
    alpha = 0.5
  ) +
  theme_bw()
```


```{r}
lcmsgcms <- lcms |>
  rename(compound = metabolite_name) |>
  mutate(instrument = "LCMS") |>
  bind_rows(gcms |> mutate(instrument = "GCMS")) |>
  select(
    sample_id,
    site,
    condition,
    dt,
    rep,
    compound,
    peak_height,
    instrument
  ) |>
  filter(!is.na(compound), !is.na(condition), !is.na(site)) |>
  # Relative abundance within each instrument x sample
  group_by(instrument, sample_id) |>
  mutate(
    relative_abundance = peak_height / sum(peak_height, na.rm = TRUE)
  ) |>
  ungroup()
```

```{r}
# CLR transform to handle compositional data
# Zero imputation: replace 0/NA with half the minimum non-zero value per instrument x sample
lcmsgcms_clr <- lcmsgcms |>
  group_by(instrument, sample_id) |>
  mutate(
    ra_imputed = if_else(
      relative_abundance == 0 | is.na(relative_abundance),
      min(relative_abundance[relative_abundance > 0], na.rm = TRUE) / 2,
      relative_abundance
    ),
    clr = log(ra_imputed) - mean(log(ra_imputed), na.rm = TRUE)
  ) |>
  ungroup()
```

```{r}
# Diagnose duplicates before pivoting
lcmsgcms_clr |>
  mutate(feature = paste(instrument, compound, sep = "__")) |>
  summarise(n = n(), .by = c(sample_id, site, condition, dt, rep, feature)) |>
  filter(n > 1L) |>
  arrange(desc(n)) |>
  print(n = 20)
```

```{r}
# Pivot to wide matrix: one row per sample, one column per instrument x compound
pca_wide <- lcmsgcms_clr |>
  mutate(feature = paste(instrument, compound, sep = "__")) |>
  select(sample_id, site, condition, dt, rep, feature, clr) |>
  pivot_wider(names_from = feature, values_from = clr)

# Separate metadata from numeric matrix
pca_meta <- pca_wide |> select(sample_id, site, condition, dt, rep)
pca_mat <- pca_wide |>
  select(-sample_id, -site, -condition, -dt, -rep) |>
  as.matrix()
rownames(pca_mat) <- pca_meta$sample_id

# Block scale: divide each instrument's block by its RMS so both contribute equally
rms <- function(x) sqrt(mean(x^2, na.rm = TRUE))
lcms_cols <- grep("^LCMS__", colnames(pca_mat))
gcms_cols <- grep("^GCMS__", colnames(pca_mat))
pca_mat[, lcms_cols] <- pca_mat[, lcms_cols] / rms(pca_mat[, lcms_cols])
pca_mat[, gcms_cols] <- pca_mat[, gcms_cols] / rms(pca_mat[, gcms_cols])

# Run PCA (center only — CLR + block scaling already handled variance structure)
pca_result <- prcomp(pca_mat, center = TRUE, scale. = FALSE)
summary(pca_result)
```

```{r}
# Extract scores and rejoin metadata for plotting
pca_scores <- pca_result$x |>
  as.data.frame() |>
  rownames_to_column("sample_id") |>
  left_join(pca_meta, by = "sample_id")

# Percent variance explained
pct_var <- round(summary(pca_result)$importance[2, ] * 100, 1)

metab_PCA <- ggplot(
  pca_scores,
  aes(x = PC1, y = PC2, color = condition, shape = site)
) +
  geom_point(size = 3) +
  scale_shape_manual(values = c("circle", "square")) +
  scale_color_manual(values = c("#4E80C7", "#FC9403")) +
  labs(
    x = paste0("PC1 (", pct_var["PC1"], "%)"),
    y = paste0("PC2 (", pct_var["PC2"], "%)")
  ) +
  theme_bw() +
  theme(aspect.ratio = 1)

metab_PCA

ggsave(
  metab_PCA,
  filename = "fig_output/metab_pca.pdf",
  height = 5,
  width = 6
)
```

# Top 10 compounds driving each PC

```{r}
# Extract loadings from the prcomp rotation matrix
loadings <- pca_result$rotation |>
  as.data.frame() |>
  rownames_to_column("feature")

# Top 10 by absolute loading for PC1 and PC2
top_pc1 <- loadings |>
  select(feature, PC1) |>
  mutate(abs_loading = abs(PC1)) |>
  slice_max(abs_loading, n = 10) |>
  arrange(desc(abs_loading))

top_pc2 <- loadings |>
  select(feature, PC2) |>
  mutate(abs_loading = abs(PC2)) |>
  slice_max(abs_loading, n = 10) |>
  arrange(desc(abs_loading))

print("Top 10 compounds driving PC1:")
print(top_pc1)

print("Top 10 compounds driving PC2:")
print(top_pc2)
```

```{r}
p_loadings <- bind_rows(
  top_pc1 |> mutate(PC = "PC1", loading = PC1),
  top_pc2 |> mutate(PC = "PC2", loading = PC2)
) |>
  mutate(feature = fct_reorder(feature, abs_loading)) |>
  ggplot(aes(x = loading, y = feature, fill = loading > 0)) +
  geom_col() +
  facet_wrap(vars(PC), scales = "free") +
  scale_fill_manual(values = c("#7eb4df", "#f77b59"), guide = "none") +
  labs(x = "Loading", y = "") +
  theme_bw()

p_loadings

ggsave(
  plot = p_loadings,
  filename = "fig_output/metab_pca_loadings.pdf",
  height = 6,
  width = 10
)
```