---
title: "04_IRMS Processing"
format: html
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(ggbiplot)
library(gt)
library(tidyplots)
```

# Load GCMS Data


```{r}
gcms <- readxl::read_excel(
  path = "data/metabolomics/mx 884963_Tristan Caro-Karthikeyan_soil extract_primary metabolism_GC-TOF MS_08-2025 submit_tidy.xlsx",
  sheet = "submit",
  skip = 8
) |>
  pivot_longer(
    cols = -c(
      `BinBase name`,
      ret.index,
      `quant mz`,
      `mass spec`,
      PubChem,
      KEGG,
      `InChI Key`
    ),
    names_to = "sample_id",
    values_to = "peak_height"
  ) |>
  mutate(
    site = case_when(
      str_detect(sample_id, "JCR") ~ "JCR",
      str_detect(sample_id, "LOM") ~ "LOM",
    ),
    condition = case_when(
      str_detect(sample_id, "ROS") ~ "Drought",
      str_detect(sample_id, "SKY") ~ "Control"
    ),
    dt = case_when(
      str_detect(sample_id, "72hr") ~ "72hr",
      str_detect(sample_id, "0hr") ~ "0hr",
      str_detect(sample_id, "24hr") ~ "24hr"
    ),
    rep = case_when(
      str_detect(sample_id, "rep1") ~ 1,
      str_detect(sample_id, "rep2") ~ 2,
      str_detect(sample_id, "rep3") ~ 3
    )
  ) |>
  select(sample_id, site, condition, dt, rep, everything()) |>
  rename(compound = `BinBase name`) |>
  # filter out unidentified compounds (represented by integers)
  filter(!str_detect(compound, "^-?\\d+$")) |>
  filter(!sample_id %in% c("BB id", "Pool 1", "Pool 2", "Pool 3", "Pool 4"))

gcms |> glimpse()
```

```{r}
# inspect all compounds
gcms |> pull(compound) |> unique() |> sort()
```

## Tile plot

```{r}
gcms |>
  ggplot(
    aes(
      x = sample_id,
      y = compound,
      fill = peak_height
    )
  ) +
  geom_tile() +
  scale_fill_viridis_c(trans = "log10") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Convert into matrix for NMDS

```{r}
gcms_nmds <- gcms |>
  select(
    sample_id,
    compound,
    peak_height,
    site,
    condition,
    dt,
    rep
  ) |>
  mutate_at(vars(site, condition, dt, rep), as.factor) |>
  filter(
    !is.na(compound),
    !is.na(site),
    !is.na(condition)
  ) |>
  pivot_wider(
    names_from = sample_id,
    values_from = peak_height
  )

gcms.wide <- gcms |>
  select(
    sample_id,
    site,
    condition,
    dt,
    rep,
    compound,
    peak_height
  ) |>
  filter(!is.na(compound)) |>
  pivot_wider(
    id_cols = c(sample_id, site, condition, dt, rep),
    names_from = compound,
    values_from = peak_height
  )

gcms.pca <- prcomp(
  gcms.wide |>
    # extract the numeric columns for PCA
    select(-c(sample_id, site, condition, dt, rep)) |>
    # missing values are replaced with 0
    mutate(across(everything(), ~ replace_na(.x, 0)))
)

plot(gcms.pca)
```

```{r}
screeplot(gcms.pca, type = "line")

p1 <- ggbiplot(
  gcms.pca,
  groups = gcms.wide$condition,
  var.axes = FALSE
) +
  theme_bw()

p2 <- ggbiplot(
  gcms.pca,
  groups = gcms.wide$site,
  shape = gcms.wide$condition,
  var.axes = FALSE
) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

cowplot::plot_grid(p1, p2, ncol = 2)
```

# GCMS Volcano plotting


```{r}
gcms_volcano <- gcms |>
  group_by(compound) |>
  dplyr::summarize(
    mean_control = mean(peak_height[condition == "Control"], na.rm = TRUE),
    mean_drought = mean(peak_height[condition == "Drought"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[condition == "Control"],
      peak_height[condition == "Drought"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_drought / mean_control),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )

# Create volcano plot
gcms_volcano |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "GC-MS Data",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))

```


```{r}
sig_metabolites <- gcms_volcano |>
  filter(significant == "Significant") |>
  arrange(p_value, log2_fc) |>
  mutate(source = "GCMS")

sig_metabolites |>
  select(compound, p_value, log2_fc) |>
  gt() |>
  data_color(
    columns = c(log2_fc),
    palette = "Blues"
  ) |>
  data_color(columns = c(p_value), palette = "Reds", reverse = TRUE) |>
  fmt_scientific()
```

## export significant metabolite dataframe

```{r}
saveRDS(sig_metabolites, file = "cache/gcms_significant_metabolites.rds")
```

# Load in LCMS datasets and combine with GCMS dataset

Significant metabolites from LCMS
```{r}
# lcms dataset
lcms <- read_rds("cache/lcms_dataset.rds")

# LCMS significant metabolites
lcms_sig <- read_rds("cache/lcms_significant_metabolites.rds") |>
  mutate(source = "LCMS")

overlapping <- intersect(lcms_sig$compound, sig_metabolites$compound)

metab_sig_combined <- lcms_sig |>
  bind_rows(sig_metabolites) |>
  mutate(
    compound = case_when(
      compound %in% overlapping ~ str_c(compound, " (", source, ")"),
      TRUE ~ compound
    )
  )

metab_sig_combined |> arrange(compound)
```

# Examine metabolite patterns from both datasets

```{r}
lcms_volcano <- read_rds("cache/lcms_volcano.rds") |> mutate(source = "LCMS")

overlapping_all <- intersect(lcms_volcano$compound, gcms_volcano$compound)

metab_combined <- lcms_volcano |>
  bind_rows(gcms_volcano |> mutate(source = "GCMS")) |>
  mutate(
    compound = case_when(
      compound %in% overlapping_all ~ str_c(compound, " (", source, ")"),
      TRUE ~ compound
    )
  ) |>
  arrange(compound)

p_metab_combined <- metab_combined |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant,
      label = compound
    )
  ) +
  geom_point(alpha = 0.5, shape = 21) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2)) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_metab_combined
```

```{r}
metab_combined |>
  writexl::write_xlsx(path = "data_output/metab_combined_volcano.xlsx")

metabolite_manual_annotations <- readxl::read_excel(
  "data/metabolomics/metabolite_manual_annotations.xlsx"
)

meta_annot <- metab_combined |>
  left_join(metabolite_manual_annotations, by = "compound")
```

```{r}
p_osmo_sig <- meta_annot |>
  filter(!is.na(annotation)) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = annotation,
      label = compound
    )
  ) +
  geom_point(alpha = 0.5, shape = 21) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_fill_manual(values = c("orange", "blue")) +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2)) +
  ggrepel::geom_text_repel(
    data = function(df) {
      df |>
        filter(
          annotation == "osmo",
          neg_log10_p > -log10(0.05) & abs(log2_fc) > 1
        )
    },
    color = "black",
    size = 3,
    force = 0.5,
    nudge_y = 0.7,
    nudge_x = 0.8,
    box.padding = 0.5,
    segment.color = "black"
  ) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_osmo_sig

p_depol_sig <- meta_annot |>
  filter(!is.na(annotation)) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = annotation,
      label = compound
    )
  ) +
  geom_point(alpha = 0.5, shape = 21) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_fill_manual(values = c("orange", "blue")) +
  coord_cartesian(xlim = c(-3, 7), ylim = c(0, 8.2)) +
  ggrepel::geom_text_repel(
    data = function(df) {
      df |>
        filter(
          annotation == "depol",
          neg_log10_p > -log10(0.05) & abs(log2_fc) > 1
        )
    },
    color = "black",
    size = 3,
    force = 0.5,
    nudge_y = 1,
    nudge_x = 0,
    box.padding = 1,
    segment.color = "black"
  ) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
p_depol_sig
```

```{r}
p_combined <- cowplot::plot_grid(
  p_metab_combined,
  p_osmo_sig,
  p_depol_sig,
  ncol = 3
)
p_combined

cowplot::save_plot(
  plot = p_combined,
  filename = "fig_output/metabolomics_combined_volcano.pdf",
  base_height = 4,
  base_width = 12
)
```

# Analyze time series data

```{r}
lcms <- read_rds("cache/lcms_full_data.rds")
```


```{r}
annotated_metabolites <- meta_annot |>
  select(compound, annotation) |>
  filter(!is.na(annotation)) |>
  mutate(
    compound = case_when(
      str_detect(compound, "\\(LCMS\\)$") ~ str_remove(
        compound,
        " \\(LCMS\\)$"
      ),
      str_detect(compound, "\\(GCMS\\)$") ~ str_remove(
        compound,
        " \\(GCMS\\)$"
      ),
      TRUE ~ compound
    )
  )

lcms_dt <- annotated_metabolites |>
  left_join(lcms |> rename(compound = metabolite_name), by = "compound")

gcms_dt <- gcms |>
  left_join(annotated_metabolites, by = "compound")
```


```{r}
# define a function for lots of plotting
tidyplot_metabolite_lcms <- function(compound_choice) {
  df <- lcms_dt |>
    mutate(
      site = case_when(
        str_detect(sample_id, "JCR") ~ "Cattle Pasture",
        str_detect(sample_id, "LOM") ~ "Grassland",
      )
    ) |>
    filter(
      !is.na(peak_height),
      !is.na(dt),
      !is.na(site),
      compound == compound_choice
    )

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  if (!any(is.finite(df$peak_height))) {
    return(invisible(NULL))
  }

  df <- df |>
    group_by(site) |>
    mutate(site_max = max(peak_height, na.rm = TRUE)) |>
    ungroup() |>
    filter(is.finite(site_max) & site_max > 0) |>
    group_by(site) |>
    mutate(normalized = peak_height / site_max) |>
    ungroup()

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  df |>
    tidyplot(x = dt, y = normalized, color = condition) |>
    add_mean_bar(alpha = 0.4) |>
    add_sem_errorbar() |>
    add(labs(
      x = "Time after rewetting",
      y = "Normalized Peak Height",
      caption = compound_choice
    )) |>
    add(scale_y_continuous(labels = scales::label_percent(), expand = FALSE)) |>
    adjust_size(height = 60, width = 60) |>
    remove_legend() |>
    split_plot(by = site) |>
    save_plot(
      filename = paste0(
        "fig_output/metabolite_plots/",
        compound_choice,
        "_lcms.png"
      )
    )
}

tidyplot_metabolite_gcms <- function(compound_choice) {
  df <- gcms_dt |>
    mutate(
      site = case_when(
        str_detect(sample_id, "JCR") ~ "Cattle Pasture",
        str_detect(sample_id, "LOM") ~ "Grassland",
      )
    ) |>
    filter(
      !is.na(peak_height),
      !is.na(dt),
      !is.na(site),
      compound == compound_choice
    )

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  if (!any(is.finite(df$peak_height))) {
    return(invisible(NULL))
  }

  df <- df |>
    group_by(site) |>
    mutate(site_max = max(peak_height, na.rm = TRUE)) |>
    ungroup() |>
    filter(is.finite(site_max) & site_max > 0) |>
    group_by(site) |>
    mutate(normalized = peak_height / site_max) |>
    ungroup()

  if (nrow(df) == 0) {
    return(invisible(NULL))
  }

  df |>
    tidyplot(x = dt, y = normalized, color = condition) |>
    add_mean_bar(alpha = 0.4) |>
    add_sem_errorbar() |>
    add(labs(
      x = "Time after rewetting",
      y = "Normalized Peak Height",
      caption = compound_choice
    )) |>
    add(scale_y_continuous(labels = scales::label_percent(), expand = FALSE)) |>
    adjust_size(height = 60, width = 60) |>
    remove_legend() |>
    split_plot(by = site) |>
    save_plot(
      filename = paste0(
        "fig_output/metabolite_plots/",
        compound_choice,
        "_gcms.png"
      )
    )
}

# uncomment to re-render all bar plots
# for (compound in annotated_metabolites$compound) {
#   tidyplot_metabolite_gcms(compound)
#   tidyplot_metabolite_lcms(compound)
# }
```


```{r}
combined_dt <- gcms_dt |>
  mutate(type = "GCMS") |>
  bind_rows(lcms_dt |> mutate(type = "LCMS")) |>
  select(
    -c(
      PubChem,
      KEGG,
      identifier,
      inchikey,
      m_z,
      rt,
      esi_mode,
      `mass spec`,
      `quant mz`,
      ret.index
    )
  ) |>
  mutate(compound = glue::glue("{compound} ({type})")) |>
  group_by(type, site, condition, compound, dt) |>
  # take mean of replicates at each time point, for each compound, for each site, for each treatment, for each method:
  dplyr::summarize(
    mean_peak_height = mean(peak_height, na.rm = TRUE)
  ) |>
  ungroup() |>
  filter(!is.na(site), !is.na(condition)) |>
  pivot_wider(
    names_from = dt,
    values_from = mean_peak_height
  ) |>
  mutate(
    percent_change_24hr = (`24hr` - `0hr`) / `0hr` * 100,
    percent_change_72hr = (`72hr` - `0hr`) / `0hr` * 100,
    fold_change_24hr = `24hr` / `0hr`,
    fold_change_72hr = `72hr` / `0hr`
  )

combined_dt |> glimpse()

combined_dt |>
  select(
    type,
    site,
    condition,
    compound,
    fold_change_24hr,
    fold_change_72hr,
    percent_change_24hr,
    percent_change_72hr
  ) |>
  arrange(percent_change_72hr) |>
  gt::gt() |>
  gt::data_color(
    columns = percent_change_72hr,
    method = "numeric",
    palette = "viridis"
  ) |>
  gt::fmt_number(decimals = 1)
```

## Analyze timeseries data with volcano plots

```{r}
lcms_volcano_dt <- lcms |>
  group_by(metabolite_name) |>
  # look only at 0hr vs 72hr data
  filter(dt != "24hr") |>
  dplyr::summarize(
    # take average of replicate measurements
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[dt == "0hr"],
      peak_height[dt == "72hr"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_72hr / mean_0hr),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )


# Create volcano plot
lcms_volcano_dt |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "LC-MS Data: 0hr vs 72hr post-rewetting",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))

```

## do the same with GCMS data


```{r}
gcms_volcano_dt <- gcms |>
  group_by(site, compound) |>
  # look only at 0hr vs 72hr data
  filter(dt != "24hr") |>
  dplyr::summarize(
    # take average of replicate measurements
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    # T-test p value
    p_value = t.test(
      peak_height[dt == "0hr"],
      peak_height[dt == "72hr"]
    )$p.value, # pull the p-value
    .groups = "drop"
  ) |>
  # calculate log2 fold change and -log10 pvalue here
  mutate(
    log2_fc = log2(mean_72hr / mean_0hr),
    neg_log10_p = -log10(p_value),
    # is it significant?
    significant = case_when(
      abs(log2_fc) > 1 & p_value < 0.05 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )


# Create volcano plot
gcms_volcano_dt |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  ggrepel::geom_text_repel(
    data = function(df) df |> filter(significant == "Significant"),
    aes(label = compound),
    color = "black"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  facet_wrap(vars(site)) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    title = "GC-MS Data: 0hr vs 72hr post-rewetting",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "None", plot.title = element_text(hjust = 0.5))
```

## combined

```{r}
gcms_volcano_dt |>
  bind_rows(lcms_volcano_dt |> rename(compound = metabolite_name)) |>
  ggplot(
    aes(
      x = log2_fc,
      y = neg_log10_p,
      fill = significant
    )
  ) +
  geom_point(shape = 21, alpha = 0.5) +
  scale_fill_manual(
    values = c("Not significant" = "gray", "Significant" = "red")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(
    x = "Log2 Fold Change Following Rewetting",
    y = "-Log10 P-Value",
    title = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "None",
    plot.title = element_text(hjust = 0.5)
  )
```

## plot relative increase or decrease


```{r}
lcms_decrease <- lcms |>
  rename(compound = metabolite_name) |>
  filter(condition == "Drought") |>
  left_join(metabolite_manual_annotations, by = "compound") |>
  mutate(
    annotation = case_when(annotation == "osmo" ~ "Osmolyte", TRUE ~ NA)
  ) |>
  filter(!is.na(site), !is.na(compound)) |>
  group_by(site, compound) |>
  mutate(
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_24hr = mean(peak_height[dt == "24hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    rel_change_72hr = (mean_72hr - mean_0hr) / mean_0hr,
    rel_change_24hr = (mean_24hr - mean_0hr) / mean_0hr
  ) |>
  ungroup() |>
  group_by(site) |>
  mutate(
    compound = fct_reorder(compound, rel_change_72hr, .desc = TRUE)
  )

p_lcms_decrease_JCR <- lcms_decrease |>
  filter(site == "JCR") |>
  ggplot(
    aes(
      x = rel_change_72hr,
      y = compound,
      color = annotation
    )
  ) +
  geom_text(
    aes(
      x = case_when(
        rel_change_72hr < 0 ~ 0.1,
        rel_change_72hr > 0 ~ -0.1
      ),
      label = compound,
      hjust = case_when(
        rel_change_72hr < 0 ~ 0,
        rel_change_72hr > 0 ~ 1
      )
    ),
    size = 1,
  ) +
  geom_segment(
    aes(xend = 0, yend = compound),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_point(
    size = 0.5
  ) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_color_manual(values = c("#fc0362", "gray")) +
  facet_wrap(vars(site)) +
  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    x = "Relative Change",
    y = ""
  ) +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
p_lcms_decrease_JCR

ggsave(
  "fig_output/JCR_lcms_rel_change_72hr.pdf",
  plot = p_lcms_decrease_JCR,
  height = 11,
  width = 8.5
)
```


```{r}
lcms_decrease <- lcms |>
  rename(compound = metabolite_name) |>
  filter(condition == "Drought", site == "LOM") |>
  left_join(metabolite_manual_annotations, by = "compound") |>
  mutate(
    annotation = case_when(annotation == "osmo" ~ "Osmolyte", TRUE ~ NA)
  ) |>
  filter(!is.na(site), !is.na(compound)) |>
  group_by(site, compound) |>
  mutate(
    mean_0hr = mean(peak_height[dt == "0hr"], na.rm = TRUE),
    mean_24hr = mean(peak_height[dt == "24hr"], na.rm = TRUE),
    mean_72hr = mean(peak_height[dt == "72hr"], na.rm = TRUE),
    rel_change_72hr = (mean_72hr - mean_0hr) / mean_0hr,
    rel_change_24hr = (mean_24hr - mean_0hr) / mean_0hr
  ) |>
  ungroup() |>
  group_by(site) |>
  mutate(
    compound = fct_reorder(compound, rel_change_72hr, .desc = TRUE)
  )

p_lcms_decrease_LOM <- lcms_decrease |>
  filter(site == "LOM") |>
  ggplot(
    aes(
      x = rel_change_72hr,
      y = compound,
      color = annotation
    )
  ) +
  geom_text(
    aes(
      x = case_when(
        rel_change_72hr < 0 ~ 0.1,
        rel_change_72hr > 0 ~ -0.1
      ),
      label = compound,
      hjust = case_when(
        rel_change_72hr < 0 ~ 0,
        rel_change_72hr > 0 ~ 1
      )
    ),
    size = 1,
  ) +
  geom_segment(
    aes(xend = 0, yend = compound),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_point(
    size = 0.5
  ) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_color_manual(values = c("#fc0362", "gray")) +
  facet_wrap(vars(site)) +
  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    x = "Relative Change",
    y = ""
  ) +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
p_lcms_decrease_LOM

ggsave(
  "fig_output/LOM_lcms_rel_change_72hr.pdf",
  plot = p_lcms_decrease_LOM,
  height = 11,
  width = 8.5
)
```