---
title: "Bulk IRMS DNA SIP"
format: html
editor_options: 
  chunk_output_type: console
---

## Setup

```{r}
# libraries
library(tidyverse)
library(scales)

# source
source("source/calculate_mu.R")
source("source/d18O_VSMOW_to_F18O.R")
source("source/calculate_sigma_mu.R")
```

```{r}
dna_sip <- readxl::read_excel(
  "data/DNA-SIP/CaroT 250805 OXS Caro_18O_DNA_SIP P263293.xlsx",
  sheet = 2
) |>
  janitor::clean_names() |>
  select(-c(comment, tray_name, well_id, type_of_material, analysis_number)) |>
  mutate(
    control = if_else(
      str_detect(sample_id, "SUCR") | str_detect(sample_id, "MQ"),
      TRUE,
      FALSE
    ),
    site = case_when(
      str_detect(sample_id, "JR") ~ "JCR",
      str_detect(sample_id, "LR") ~ "LOM",
      str_detect(sample_id, "JS") ~ "JCR",
      str_detect(sample_id, "LS") ~ "LOM",
      TRUE ~ NA
    ),
    condition = case_when(
      str_detect(sample_id, "JR") ~ "Drought",
      str_detect(sample_id, "LR") ~ "Drought",
      str_detect(sample_id, "JS") ~ "Control",
      str_detect(sample_id, "LS") ~ "Control",
      TRUE ~ NA
    ),
    dt_hr = case_when(
      str_detect(sample_id, "0") ~ 0,
      str_detect(sample_id, "24") ~ 24,
      str_detect(sample_id, "72") ~ 72,
      TRUE ~ NA
    ),
    rep = case_when(
      str_detect(sample_id, "-1") ~ 1,
      str_detect(sample_id, "-2") ~ 2,
      str_detect(sample_id, "-3") ~ 3,
      TRUE ~ NA
    )
  )
glimpse(dna_sip)
```

How much oxygen was actually analyzed?

```{r}
total_o_mean <- dna_sip |> pull(total_o_mg) |> mean()
total_o_sd <- dna_sip |> pull(total_o_mg) |> sd()

message(
  "Mean total O analyzed: ",
  round(total_o_mean, 2),
  " ug +/- ",
  round(total_o_sd, 2),
  " ug"
)
```

```{r}
controls <- dna_sip |> filter(control)

controls_summary <- controls |>
  summarise(
    mean_d18o = mean(d18ovsmow),
    sd_d18o = sd(d18ovsmow)
  )

p_controls_summary <- controls |>
  ggplot(aes(x = sample_id, y = d18ovsmow)) +
  geom_rect(
    ymin = controls_summary$mean_d18o - controls_summary$sd_d18o,
    ymax = controls_summary$mean_d18o + controls_summary$sd_d18o,
    xmin = -Inf,
    xmax = Inf,
    fill = "darkblue",
    alpha = 0.01
  ) +
  geom_point() +
  geom_hline(
    yintercept = controls_summary$mean_d18o,
    linetype = "dotted",
    color = "darkblue"
  ) +
  scale_y_continuous(
    labels = label_number(style_positive = "plus", accuracy = 0.1)
  ) +
  theme_classic() +
  labs(
    x = "Sample ID",
    y = latex2exp::TeX("$\\delta^{18}O_{VSMOW}$ (‰)")
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey", linewidth = 0.1)
  )
p_controls_summary

controls_d18O_mean <- controls |> pull(d18ovsmow) |> mean()
controls_F18O_mean <- d18O_VSMOW_to_F18O(controls_d18O_mean, permil = TRUE)
message(
  "Mean d18O of controls is ",
  round(controls_d18O_mean, 2),
  " permil vs. VSMOW or ",
  controls_F18O_mean,
  " atom fraction"
)
```

run mass balance for sucrose and DNA oxygen contributions (remember, we diluted the DNA-O 1:100 with sucrose-O prior to analysis in order to achieve sufficient mass for IRMS analysis!)

mass balance calculated as:

$$
\begin{aligned}
m_T F_T = m_1 F_1 + m_2 F_2 \\
F_1 = \frac{m_T F_T - m_2 F_2}{m_1}
\end{aligned}
$$

where T is measured, 1 is pool 1 (DNA), and 2 is pool 2 (sucrose) m1 = 1 ug m2 = 100 ug F1 = unknown F2 = mean of sucrose controls (30.04 permil = 0.00206... 18F ) m_T = 101 ug F_T = measured 18O

```{r}
mass_balance_DNA_18O <- function(m_T, F_T, m_DNA, m_sucr, F_sucr) {
  F_DNA = (m_T * F_T - m_sucr * F_sucr) / m_DNA
  return(F_DNA)
}

mass_balance_necromass <- function(m_T, F_T, m_DNA, m_necro, F_necro) {
  F_DNA = F_DNA = (m_T * F_T - m_necro * F_necro) / m_DNA
  return(F_DNA)
}

# mass_balance_necromass(m_T = 1, F_T = 0.03, m_DNA = 0.5, m_necro = 0.5, F_necro = controls_F18O_mean)

# example:
# mass_balance_DNA_18O(
#   m_T = 101,
#   F_T = 0.0021,
#   m_DNA = 1,
#   m_sucr = 100,
#   F_sucr = controls_F18O_mean
# )

dna_sip_summarized <- dna_sip |>
  filter(!control) |>
  # summarize triplicates
  group_by(condition, site, dt_hr) |>
  summarise(
    mean_d18ovsmow = mean(d18ovsmow),
    sd_d18ovsmow = sd(d18ovsmow)
  ) |>
  mutate(
    # convert delta notation to fractional abundance (F)
    F18O = d18O_VSMOW_to_F18O(mean_d18ovsmow, permil = TRUE),
    F18O_t0 = d18O_VSMOW_to_F18O(
      controls_summary$mean_d18o,
      permil = TRUE
    ),
    F18O_sd = d18O_VSMOW_to_F18O(
      mean_d18ovsmow + sd_d18ovsmow,
      permil = TRUE
    ) -
      d18O_VSMOW_to_F18O(mean_d18ovsmow - sd_d18ovsmow, permil = TRUE),
    F18O_t0_sd = d18O_VSMOW_to_F18O(
      controls_summary$mean_d18o + controls_summary$sd_d18o,
      permil = TRUE
    ) -
      d18O_VSMOW_to_F18O(
        controls_summary$mean_d18o - controls_summary$sd_d18o,
        permil = TRUE
      ),
    # mass balance calculations:
    F18O_corr = mass_balance_DNA_18O(
      m_T = 101,
      F_T = F18O,
      m_DNA = 1,
      m_sucr = 100,
      F_sucr = controls_F18O_mean
    ),
    # correct for necromass-derived DNA (estimated at 0.4 (40%) of total DNA)
    F18O_necro_corr = mass_balance_necromass(
      m_T = 1,
      F_T = F18O_corr,
      m_DNA = 0.6,
      m_necro = 0.4,
      F_necro = controls_F18O_mean
    )
  )

p_d18O_raw <- dna_sip_summarized |>
  mutate(condition = factor(condition, levels = c("Drought", "Control"))) |>
  ggplot(
    aes(
      x = dt_hr,
      y = mean_d18ovsmow,
      shape = condition
    )
  ) +
  geom_line() +
  geom_hline(
    yintercept = controls_summary$mean_d18o,
    color = "darkblue",
    linetype = "dotted"
  ) +
  geom_pointrange(
    aes(
      fill = condition,
      ymin = mean_d18ovsmow - sd_d18ovsmow,
      ymax = mean_d18ovsmow + sd_d18ovsmow
    ),
    color = "black",
    shape = 21
  ) +
  facet_grid(site ~ condition) +
  scale_y_continuous(labels = label_number(style_positive = "plus")) +
  scale_fill_manual(values = c("#ffac47", "#4e80c7")) +
  theme_minimal() +
  labs(
    x = "Incubation Time (hr)",
    y = latex2exp::TeX("$\\delta^{18}O_{VSMOW}$ (‰)")
  ) +
  theme(
    panel.grid.minor.x = element_blank()
  )
p_d18O_raw


p_F18O_raw <- dna_sip_summarized |>
  mutate(condition = factor(condition, levels = c("Drought", "Control"))) |>
  ggplot(
    aes(
      x = dt_hr,
      y = F18O_necro_corr,
      fill = condition
    )
  ) +
  geom_line() +
  # plot negative controls:
  geom_hline(
    yintercept = d18O_VSMOW_to_F18O(controls_summary$mean_d18o, permil = TRUE),
    color = "darkblue",
    linetype = "dotted"
  ) +
  geom_pointrange(
    aes(ymin = F18O_necro_corr - F18O_sd, ymax = F18O_necro_corr + F18O_sd),
    shape = 21
  ) +
  facet_grid(site ~ condition) +
  scale_fill_manual(values = c("#ffac47", "#4e80c7")) +
  theme_minimal() +
  labs(
    x = "Incubation Time (hr)",
    y = latex2exp::TeX("$^{18}F$")
  ) +
  theme(
    panel.grid.minor.x = element_blank()
  )
p_F18O_raw
```

save cache

```{r}
saveRDS(dna_sip_summarized, file = "cache/dna_sip_data.rds")
```

estimate F_L based on soil water content

```{r}
gwc <- readxl::read_excel(
  "data/sample_matrix_complete.xlsx",
  sheet = "Gravimetric Water Content"
) |>
  filter(
    sample_id != "EXAMPLE"
  ) |>
  janitor::clean_names() |>
  filter(pre_or_post_rewetting == "PRE") |>
  select(
    sample_id,
    site,
    condition,
    gmc_percent
  ) |>
  mutate(
    gwc = gmc_percent / 100, # convert to fraction
    ml_water_in_soil = 0.3 * gwc, # 300mg soil
    F18O_soil_water = 0.002005, # using VSMOW as placeholder
    ml_tracer_added = 0.2, # 200uL of tracer added
    F18O_tracer = 0.20, # 20 atom % 18O
    ml_incubation_soln = ml_tracer_added + ml_water_in_soil,
    F18O_incubation = ((ml_water_in_soil * F18O_soil_water) +
      (ml_tracer_added * F18O_tracer)) /
      ml_incubation_soln,
    drought = case_when(
      condition == "ROS" ~ "Drought",
      condition == "SKY" ~ "Control"
    )
  ) |>
  select(-condition) |>
  rename(condition = drought)

gwc_mini <- gwc |>
  select(site, condition, FL = F18O_incubation)
```

```{r}

#d18O_VSMOW_to_F18O(0, permil = FALSE)

dna_sip_growth <- dna_sip_summarized |>
  left_join(gwc_mini, by = c("site", "condition")) |>
  ungroup() |>
  mutate(
    # supply parameters for growth calculation
    a_w = 0.7, # fraction of oxygen in DNA from water
    dt_d = dt_hr / 24, # convert hours to days
    mu_d = calculate_mu(
      a = a_w,
      F2 = F18O_necro_corr,
      F1 = F18O_t0,
      FL = FL,
      t2 = dt_d,
      t1 = 0
    ),
    # propagate error
    mu_d_se = calculate_sigma_mu(
      a = a_w,
      F2 = F18O_necro_corr,
      F1 = F18O_t0,
      FL = FL,
      t2 = dt_d,
      t1 = 0,
      sigma_a = 0,
      sigma_F2 = F18O_sd,
      sigma_F1 = F18O_t0_sd,
      sigma_FL = 0
    ),
    # remove infinite values
    mu_d = if_else(mu_d == Inf | mu_d == -Inf, 0, mu_d),
    mu_d_se = if_else(mu_d_se == Inf, 0, mu_d_se),
    gen_d = log(2) / mu_d,
    gen_d = if_else(gen_d == 0 | gen_d == Inf, NA, gen_d)
  )

dna_sip_growth |>
  # refactor for plotting panels in order
  mutate(condition = factor(condition, levels = c("Drought", "Control"))) |>
  # plot it!
  ggplot(
    aes(
      x = dt_hr,
      y = mu_d,
      fill = gen_d
    )
  ) +
  geom_line(linetype = "dotted") +
  geom_pointrange(
    aes(
      ymin = mu_d - mu_d_se,
      ymax = mu_d + mu_d_se
    ),
    shape = 21,
  ) +
  scale_fill_viridis_c(
    begin = 1,
    end = 0
  ) +
  facet_grid(site ~ condition) +
  theme_minimal() +
  labs(
    x = "Incubation Time (hr)",
    y = latex2exp::TeX("$\\mu_{DNA}$ (d$^{-1}$)"),
    fill = "Generation Time (d)"
  ) +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.2, "cm"),
    legend.key.width = unit(1, "cm"),
  ) +
  guides(
    fill = guide_colourbar(title.position = "left")
  )
```

```{r}
#saveRDS(dna_sip_growth, file = "cache/dna_sip_growth.rds")
```

# compare with lipid SIP growth rates

```{r}
lipid_sip_growth <- read_rds("cache/lipid_sip_growth_data.rds") |>
  filter(SIP_type != "Vapor-SIP") |>
  select(
    clean_id,
    site,
    dt_hr,
    condition = drought,
    mu_d_lipid = u_d,
    mu_d_lipid_se = s_u_d,
    gen_d_lipid = gen_d
  ) |>
  mutate(dt_d = dt_hr / 24) |>
  mutate(
    condition = case_when(
      condition == "Drought" ~ "Drought",
      condition == "No Drought" ~ "Control"
    )
  )

# join the two dfs
compare_sips <- dna_sip_growth |>
  rename(mu_d_dna = mu_d, mu_d_dna_se = mu_d_se, gen_d_dna = gen_d) |>
  left_join(lipid_sip_growth, by = c("site", "condition", "dt_hr", "dt_d")) |>
  select(
    site,
    condition,
    dt_d,
    mu_d_dna,
    mu_d_dna_se,
    mu_d_lipid,
    mu_d_lipid_se,
    gen_d_dna,
    gen_d_lipid
  )
```

```{r}
how_much_faster <- compare_sips |>
  filter(dt_d != 0) |>
  mutate(ratio = mu_d_dna / mu_d_lipid) |>
  pull(ratio) |>
  mean()

how_much_faster_min <- compare_sips |>
  filter(dt_d != 0) |>
  mutate(ratio = mu_d_dna / mu_d_lipid) |>
  pull(ratio) |>
  min()

how_much_faster_max <- compare_sips |>
  filter(dt_d != 0) |>
  mutate(ratio = mu_d_dna / mu_d_lipid) |>
  pull(ratio) |>
  max()

message(
  "DNA synthesis rates are approximately ",
  round(how_much_faster, 2),
  " times faster than lipid synthesis rates."
)

p_compare_sips <- compare_sips |>
  mutate(condition = factor(condition, levels = c("Drought", "Control"))) |>
  filter(dt_d != 0) |>
  ggplot(
    aes(
      x = mu_d_lipid,
      y = mu_d_dna
    )
  ) +
  geom_line(
    aes(
      group = interaction(site, condition)
    ),
    arrow = arrow(length = unit(0.3, "cm"))
  ) +
  #geom_point(aes(shape = as.factor(dt_d), color = condition)) +
  geom_pointrange(
    aes(
      xmin = mu_d_lipid - mu_d_lipid_se,
      xmax = mu_d_lipid + mu_d_lipid_se,
      ymin = mu_d_dna - mu_d_dna_se,
      ymax = mu_d_dna + mu_d_dna_se,
      fill = condition,
      shape = as.factor(dt_d)
    ),
    size = .5,
    color = "black"
  ) +
  geom_abline(slope = 1) +
  scale_fill_manual(values = c("Drought" = "#ffac47", "Control" = "#4e80c7")) +
  scale_shape_manual(values = c(21, 22)) +
  coord_cartesian(xlim = c(0.02, 0.6), ylim = c(0.02, 0.6), expand = FALSE) +
  theme_classic() +
  labs(
    x = latex2exp::TeX("$\\mu_{Lipid}$ (d$^{-1}$)"),
    y = latex2exp::TeX("$\\mu_{DNA}$ (d$^{-1}$)"),
    fill = "Condition"
  ) +
  theme(
    panel.border = element_rect(color = "black", linewidth = 1),
    axis.line = element_blank(),
    panel.grid = element_blank(),
    aspect.ratio = 1
  )

p_compare_sips
```

```{r}
ggsave(
  filename = "fig_output/sip_compare_a_w.pdf",
  plot = p_compare_sips,
  height = 5,
  width = 5
)
```