---
title: "singleM"
---


```{r}
library(tidyverse)
```

# read + tidy

Each column in the singleM summarise output encodes a sample as
`Caro-Soil-SIP-pooled__{sample_id}_{n}__{library}__{lane}_...`, where
`sample_id` is a 4-character code (e.g. `JR71`). The first 3 characters are
the sample code (site/condition/timepoint); the 4th character is an integer
replicate (1-3). Each sample was also sequenced across 3 lanes, so every
sample code has 9 columns (3 replicates x 3 lanes) that all collapse to one
averaged value per taxon.

The 3-character sample code itself is also structured, one letter per factor:

- 1st character: site — `J` = Pasture, `L` = Grassland
- 2nd character: condition — `R` = Drought, `S` = Control
- 3rd character: timepoint — `7` = Rewet, `P` = Not Rewet

```{r}
level_order <- c(
  "domain",
  "phylum",
  "class",
  "order",
  "family",
  "genus",
  "species"
)

read_singlem_level <- function(path) {
  level <- str_extract(basename(path), "(?<=-)[a-z]+(?=\\.tsv$)")

  read_tsv(path, show_col_types = FALSE) |>
    pivot_longer(
      cols = -taxonomy,
      names_to = "sample",
      values_to = "rel_abund"
    ) |>
    mutate(
      sample_id = str_extract(
        sample,
        "(?<=Caro-Soil-SIP-pooled__)[A-Za-z0-9]{4}"
      ),
      sample_code = str_sub(sample_id, 1, 3),
      replicate = as.integer(str_sub(sample_id, 4, 4))
    ) |>
    summarize(rel_abund = mean(rel_abund), .by = c(taxonomy, sample_code)) |>
    mutate(level = level)
}

taxonomy_long <- list.files(
  "data/singleM/summarise/",
  pattern = "\\.tsv$",
  full.names = TRUE
) |>
  map(read_singlem_level) |>
  list_rbind() |>
  mutate(
    level = factor(level, levels = level_order),
    site = case_match(
      str_sub(sample_code, 1, 1),
      "J" ~ "Pasture",
      "L" ~ "Grassland"
    ),
    condition = case_match(
      str_sub(sample_code, 2, 2),
      "R" ~ "Drought",
      "S" ~ "Control"
    ),
    timepoint = case_match(
      str_sub(sample_code, 3, 3),
      "7" ~ "Rewet",
      "P" ~ "Not Rewet"
    ),
    site = factor(site, levels = c("Pasture", "Grassland")),
    condition = factor(condition, levels = c("Control", "Drought")),
    timepoint = factor(timepoint, levels = c("Not Rewet", "Rewet"))
  )

taxonomy_long |> glimpse()
```

# bar charts

Each level has too many taxa to color individually, so each chart keeps the
top 15 taxa by mean relative abundance and folds everything else into "Other".
Charts are faceted by site (rows) and condition (columns), with timepoint on
the x-axis.

```{r}
#| fig-width: 8
#| fig-height: 5

clean_taxon_label <- function(x) {
  # taxonomy strings are full lineages like "Root; d__Bacteria; p__Actinomycetota";
  # for legends we only want the deepest rank's name (e.g. "Actinomycetota")
  cleaned <- x |>
    str_extract("[^;]+$") |>
    str_trim() |>
    str_remove("^[a-z]__")
  if_else(cleaned == "" | is.na(cleaned), x, cleaned)
}

plot_taxonomy_bar <- function(data, level_name, n_top = 15) {
  # 15 hues interleaved across two lightness bands (needed for enough pairwise
  # separation at n = 15) plus a desaturated gray for "Other"
  palette <- c(
    "#E88D88",
    "#AE5600",
    "#C8A049",
    "#6E7500",
    "#8FB248",
    "#008800",
    "#1CBC84",
    "#0084B1",
    "#00BCC0",
    "#0077D1",
    "#60ADE6",
    "#AC30CC",
    "#BF94E8",
    "#CB217E",
    "#E885C4",
    "#898781"
  )

  level_data <- data |> filter(level == level_name)

  top_taxa <- level_data |>
    summarize(mean_abund = mean(rel_abund), .by = taxonomy) |>
    slice_max(mean_abund, n = n_top) |>
    pull(taxonomy)

  plot_data <- level_data |>
    mutate(taxonomy = if_else(taxonomy %in% top_taxa, taxonomy, "Other")) |>
    summarize(
      rel_abund = sum(rel_abund),
      .by = c(site, condition, timepoint, taxonomy)
    ) |>
    mutate(
      taxonomy = clean_taxon_label(taxonomy),
      taxonomy = fct_reorder(taxonomy, rel_abund, .fun = sum, .desc = TRUE),
      taxonomy = fct_relevel(taxonomy, "Other", after = Inf)
    )

  fill_colors <- palette[seq_len(nlevels(plot_data$taxonomy))]
  names(fill_colors) <- levels(plot_data$taxonomy)

  ggplot(plot_data, aes(x = timepoint, y = rel_abund, fill = taxonomy)) +
    geom_col(width = 0.8, color = "white", linewidth = 0.2) +
    scale_fill_manual(values = fill_colors) +
    facet_grid(rows = vars(site), cols = vars(condition)) +
    labs(
      title = str_to_title(level_name),
      x = NULL,
      y = "Relative abundance (%)",
      fill = str_to_title(level_name)
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.x = element_blank(),
      #strip.background = element_rect(fill = "grey90", color = NA)
    )
}

walk(level_order, function(lvl) {
  print(plot_taxonomy_bar(taxonomy_long, lvl))
})
```
