---
title: "gRodon"
format: html
editor_options: 
  chunk_output_type: console
highlight-style: github
---

## Setup

```{r}
library(tidyverse)
```

### Import from cache

We start by importing the growth data we generated in the lipid SIP analysis. The column we're interested in is `u_d` or $\mu_d$ which is the growth rate in units of inverse days.

```{r}
growth <- read_rds("cache/lipid_sip_growth_data.rds")
growth |> glimpse()
```

## Calculations

Hoehler et al. (2023) find that there exists a linear relationship between an organism's biomass turnover (here, it's apparent "growth" rate) and its cellular energy flux, expressed as mass-specific power (W/g C).

We consider the mass specific carbon turnover rate $\mu^*$ in mass carbon biomass synthesized $(C_{biosynthesis})$ per unit time, or $\mu^* = ((C_{biosynthesis} \times s^{-1} \times (C_{biomass})^{-1}$

MSP is related to the biomass yield $(Y^* = C_{biosynthesis} /J)$ of an organism by:

$$
MSP =\frac{\mu^*}{Y^*}
$$

Hoehler et al. find that biomass yield (Y\*) for aerobic organisms is related to their mass-specific carbon turnover rate $(yr^{-1})$ by the relationship:

$$
Y^* = 0.019 \pm 0.008 \: g \: C \: kJ^{-1}
$$

Lipid-SIP-derived estimates of bulk community biomass turnover therefore can constrain the bulk soil MSP:

$$
MSP = \frac{\mu^* \: (yr)^{-1}}{0.019 \: gC \: (kJ)^{-1}} = \frac{\mu^* \: (yr)^{-1}}{1.9\times10^{-5} \: gC \: (J)^{-1}}
$$

```{r}
growth <- growth |>
  mutate(
    u_yr = u_d * 365, # convert from per day to per year
    u_sec = u_d / 86400, # convert from per day to per second
    MSP_w_gc = u_sec / 1.9e-5 # convert to MSP (W/g C)
  )

```

```{r}
# visualize the distribution in u_yr
growth |>
  ggplot(aes(x = u_yr, fill = SIP_type)) +
  geom_histogram(bins = 20, color = "white") +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = latex2exp::TeX("$\\mu^*$ (yr$^{-1}$)"),
    y = "Count"
  )
```

```{r}
growth |>
  ggplot(
    aes(
      y = u_yr,
      x = MSP_w_gc,
      color = SIP_type
    )
  ) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    y = latex2exp::TeX("$\\mu^*$ (yr$^{-1}$)"),
    x = latex2exp::TeX("Mass-specific power (W/g C)")
  )
```

Once we have our mass specific power estimates, we can estimate the respiration rate that is required to sustain this energetic flux. First, we must calculate total energy flux (without units of mass) estimated from the total biomass in the system.

Let's use the actual lipid biomass data estimates from the system:

```{r}
# read in the lipid data
lipid_biomass <- read_rds("cache/lipid_abundances.rds") |> ungroup()

lipid_biomass_summary <- lipid_biomass |>
  group_by(site) |>
  summarise(
    total_ug_per_g_soil = mean(total_ug_per_g_soil)
  ) |>
  mutate(site = case_when(site == "Loma Ridge" ~ "LOM", TRUE ~ site))
```

We conservatively assume that lipids are 9% of total cell biomass (Neidhardt et al. 1987)

```{r}
growth_w_biomass <- growth |>
  left_join(
    lipid_biomass_summary |> select(site, total_ug_per_g_soil),
    by = "site"
  ) |>
  rename(lipid_biomass_ug_per_g_soil = total_ug_per_g_soil) |>
  mutate(
    cells_per_g_soil = case_when(
      site == "JCR" ~ 4e8,
      site == "LOM" ~ 1e8
    ),
    single_cell_biomass_fg = 100,
    cell_biomass_fg_per_g_soil = cells_per_g_soil * single_cell_biomass_fg,
    cell_biomass_g_per_g_soil = cell_biomass_fg_per_g_soil * 1e-15
  )

```

From Hoehler et al., we apply an estimate of the $\Delta G_r$ of soil aerobic heterotrophic respiration of organic matter:

$$
\Delta G_r = -4.83 \times 10^{-5} \: J \: (mol \: C)^{-1}
$$

for the chemical equation:

$$
CH_2O + O_2 \rightarrow CO_2 + H_2O
$$

Total power (P) is calculated as the MSP multiplied by total biomass:

$$
P = W \: (gC)^{-1} \cdot (gC) = W
$$

therefore, we can calculate the mol C respired per g of soil per second:

$$
mol \: C \:(sec)^{-1} = \frac{P}{\Delta G_r} = \frac{J \: (sec)^{-1}}{-4.83\times 10^{5} \: J \: (mol \: C)^{-1}}
$$

Our respirometry measurements (CO2 fluxes) are reported in units of µmol per square meter per second. To get to equal units, we consider a 10cm x 1m x 1m slab of soil with an assumed bulk density of 1.5 g/mL:

$$
\mu mol \: C \: (m^2 \cdot sec)^{-1} = \mu mol \: C \: (sec \cdot g_{soil})^{-1} \times \frac{1.5 g}{cm^{3}} \times 10^5 \: cm^3
$$

```{r}
growth_w_biomass_flux <- growth_w_biomass |>
  mutate(
    deltaG_r_J_per_molC = -4.83e5, # J per mol C respired
    total_power_W_per_g_soil = MSP_w_gc * cell_biomass_g_per_g_soil,
    molC_respired_per_sec_per_g_soil = total_power_W_per_g_soil /
      abs(deltaG_r_J_per_molC), # mol C per sec per g soil
    umolC_respired_per_sec_per_g_soil = molC_respired_per_sec_per_g_soil * 1e6, # mol to umol
    umolC_respired_per_sec_per_m2_soil = umolC_respired_per_sec_per_g_soil *
      1.5 *
      1e5 # g soil to cm3 to m2 (10cm depth, 1.5 g/cm3 bulk density)
  )

```

```{r}
growth_w_biomass_flux |>
  ggplot(aes(x = umolC_respired_per_sec_per_m2_soil, fill = SIP_type)) +
  geom_histogram(bins = 20, color = "white") +
  scale_x_log10() +
  theme_minimal() +
  labs(
    y = "Count"
  )
```

This plot should look similar in shape to the growth rate plot.

```{r}
growth_w_biomass_flux |>
  ggplot(
    aes(
      y = umolC_respired_per_sec_per_m2_soil,
      x = dt_hr,
      shape = SIP_type
    )
  ) +
  geom_point() +
  facet_grid(site ~ drought) +
  theme_minimal() +
  labs(
    x = "Incubation time (hr)",
    y = latex2exp::TeX(
      "Inferred Respiration rate $\\left( \\mu mol \\, C  \\, (m^2 \\, sec \\right) ^{-1}$"
    )
  )
```

## Compare to measured CO2 fluxes

Finally, we can compare our inferred respiration rates to the measured CO2 fluxes from the soil cores. We expect that our inferred respiration rates should be lower than the measured CO2 fluxes, since some of the CO2 flux is likely derived from spontaneous, abiotic release of CO2 from soil.

```{r}
# make this tibble similar to our growth tibble
co2_flux <- read_rds("cache/co2_flux_umol_m2_sec.rds") |>
  rename(
    dt_hr = time_after_rain,
    drought = condition,
    umol_emitted_per_sec_per_m2_soil = umol_m2_sec
  ) |>
  mutate(
    site_abbrev = str_sub(site, 1, 1),
    drought_abbrev = str_sub(drought, 1, 1) |> str_to_upper(),
    SIP_type = case_when(
      dt_hr == 0 ~ "Vapor-SIP",
      dt_hr != 0 ~ "Liquid-SIP"
    ),
    SIP_type_abbrev = case_when(
      SIP_type == "Vapor-SIP" ~ "V",
      SIP_type == "Liquid-SIP" ~ "W"
    ),
    clean_id = paste0(site_abbrev, drought_abbrev, SIP_type_abbrev, dt_hr)
  ) |>
  select(
    clean_id,
    everything(),
    -c(site_abbrev, drought_abbrev, SIP_type_abbrev)
  )
co2_flux |> glimpse()

# take the mean of the vapor-SIP timepoints to get a single estimate of baseline respiration before rewetting
time_zero_respiration <- co2_flux |>
  filter(dt_hr == 0) |>
  group_by(site, drought, SIP_type) |>
  summarize(
    t0_umol_emitted_per_sec_per_m2_soil = mean(
      umol_emitted_per_sec_per_m2_soil,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    drought = case_when(
      drought == "ROS" ~ "Drought",
      drought == "sky" ~ "No Drought"
    )
  )
```

Join the two datasets and clean them up to be comparable for plotting.

```{r}
combined <- growth_w_biomass_flux |>
  filter(!dt_hr %in% c(3, 72)) |> # remove the 3hr and 72 hr timepoints because we don't have CO2 flux data for them
  select(
    clean_id,
    site,
    SIP_type,
    drought,
    dt_hr,
    umolC_respired_per_sec_per_m2_soil
  ) |>
  left_join(
    co2_flux |> select(clean_id, umol_emitted_per_sec_per_m2_soil),
    by = "clean_id"
  ) |>
  left_join(time_zero_respiration, by = c("site", "drought", "SIP_type")) |>
  mutate(
    umol_emitted_per_sec_per_m2_soil = case_when(
      is.na(umol_emitted_per_sec_per_m2_soil) ~
        t0_umol_emitted_per_sec_per_m2_soil,
      TRUE ~ umol_emitted_per_sec_per_m2_soil
    )
  ) |>
  select(-t0_umol_emitted_per_sec_per_m2_soil) |>
  pivot_longer(
    cols = c(
      umolC_respired_per_sec_per_m2_soil,
      umol_emitted_per_sec_per_m2_soil
    ),
    names_to = "type",
    values_to = "umol_C_per_sec_per_m2_soil"
  ) |>
  mutate(
    type = case_when(
      type == "umolC_respired_per_sec_per_m2_soil" ~
        "Inferred Respiration Rate",
      type == "umol_emitted_per_sec_per_m2_soil" ~ "Measured CO2 Flux"
    )
  )
```

Plot it

```{r}
combined |>
  filter(SIP_type != "Vapor-SIP") |> # only plot liquid SIP data
  ggplot(
    aes(
      y = umol_C_per_sec_per_m2_soil,
      x = as.factor(dt_hr),
      fill = type
    )
  ) +
  geom_col(
    color = "white",
    position = position_dodge(width = 0.5),
    width = 0.5
  ) +
  facet_grid(site ~ drought) +
  theme_minimal()

```