---
title: "Vapor Labeling"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

## Setup

```{r}
library(tidyverse)
library(pracma)
library(tidymodels)
library(broom)
```

```{r}
?pracma::integral
```

## Test Dataset

First, we fit a negative exponential function as described in Ingraham and Criss:

$$
^{18}F_{sw} = \: ^{18}F_T + (^{18}F_{i} - ^{18}F_T) * (e^{bt})
$$

where:

-   $^{18}F_{sw}$ is the 18O fractional abundance of soil water (sw)

-   $^{18}F_{T}$ is the 18O fractional abundance of incubation water at time T

-   ${18}F_i$ is the 18O fractional abundance of incubation water at initial time (i)

-   $b$ is a soil-specific coefficient generated by fitting nls() function in R.

```{r}
water_df <- tibble(
  t = c(0, 3, 6, 17, 48),
  y = c(65, 55, 45, 33, 22),
  sample = "sample1"
)

water_df |>
  ggplot() +
  aes(x = t, y = y) +
  geom_point() +
  theme_classic()
```

```{r}
water_fit_test <- nls(y ~ SSasymp(t, yf, y0, log_alpha), data = water_df) |>
  augment()

water_fit_test |>
  ggplot(
    aes(x = t, y = y)
  ) +
  geom_point() +
  geom_line(aes(y = .fitted), color = "red")
```

```{r}
water_df2 <- tibble(
  t = c(0, 3, 6, 17, 48),
  at18O = c(63, 50, 53, 30, 20),
  sample = "sample2"
) |>
  bind_rows(water_df)

water_tidy <- water_df2 |>
  nest(data = -sample) |> # put everything into nested dfs in a col called "data"
  mutate(
    fit = map(data, ~ nls(at18O ~ SSasymp(t, yf, y0, log_alpha), data = .x)),
    tidied = map(fit, tidy),
    augmented = map(fit, augment)
  )

water_tidy |>
  unnest(tidied) |>
  select(sample, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  mutate(alpha = exp(log_alpha))
```

```{r}
water_augmented <- water_tidy |>
  unnest(augmented)

water_augmented |>
  ggplot(aes(x = time, y = at18O, group = sample)) +
  geom_point(aes(color = sample)) +
  geom_line(aes(y = .fitted), color = "red")
```

Integrate $y(t) = y_f + (y_0 - y_f)e^{-log(\alpha)t}$

```{r}
water_fit_test <- nls(y ~ SSasymp(t, yf, y0, log_alpha), data = water_df)

params <- coef(water_fit_test)

# Define the exponential decay function
exp_decay <- function(t, yf, y0, alpha) {
  yf + (y0 - yf) * exp(-alpha * t)
}

test <- tibble(
  x = seq(0, 48, by = 1),
  y = exp_decay(
    t = x,
    yf = params['yf'],
    y0 = params['y0'],
    alpha = exp(params['log_alpha'])
  )
)

# Check if our fit looks good
test |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_point(data = water_df, aes(x = t, y = y), color = "blue")
# okay??? the fit looks fine?!?!

# Integrate the function between t1 and t2
integral <- function(t1, t2, yf, y0, alpha) {
  integrate(
    function(t) exp_decay(t, yf, y0, alpha),
    lower = t1,
    upper = t2
  )$value
}

# Example: integrate the function between t1 = 0 and t2 = 10
integral_value <- integral(
  0,
  48,
  yf = params['yf'],
  y0 = params['y0'],
  alpha = exp(params['log_alpha'])
)

# divide by the total time
integral_value / 48
```
